---
title: "Causal Machine Learning assignment"
author: "Thomas Sertijn, Bart Smets, Ilja Van Bever"
date: "`r Sys.Date()`"
output: pdf_document
header-includes:
  - \usepackage{tikz}
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(hdm)
library(grf)
library(dplyr)
library(tidyr)
library(knitr)
#load data
data(pension)
library(ggdag)
library(ggplot2)

```

```{r, echo = FALSE}
# Cross-fitting

# Reference doc:
# https://grf-labs.github.io/grf/reference/causal_forest.html
# https://grf-labs.github.io/grf/reference/regression_forest.html

set.seed(123)

variables <- c("net_tfa", "e401", "age", "inc", "fsize", "educ", 
          "marr", "twoearn", "db", "pira", "hown")


Y <- pension$net_tfa
A <- pension$e401
X <- pension[, c("age", "inc", "fsize", "educ", 
                                 "marr", "twoearn", "db", "pira", "hown")]
AX <- pension[, c("e401", "age", "inc", "fsize", "educ", 
                                 "marr", "twoearn", "db", "pira", "hown")]

K = 5 # set to 5!
n = nrow(X)
folds = sample(rep(1:K, length.out = n))

Y_X_hat = rep(NA, n)
Y_AX_hat = rep(NA, n)
A_hat = rep(NA, n)

for (k in 1:K) {
  
  train = which(folds != k)
  test  = which(folds == k)
  
  # E[Y|X]
  rf_Y_X <- regression_forest(X[train, ], Y[train])
  Y_X_hat[test] <- predict(rf_Y_X, X[test, ])$predictions
  
  # E[A|X]
  rf_A <- regression_forest(X[train, ], A[train])
  A_hat[test] <- predict(rf_A, X[test, ])$predictions
  
  # E[Y|A, L] (exercise 3)
  rf_Y_AX <- regression_forest(AX[train, ], Y[train])
  Y_AX_hat[test] <- predict(rf_Y_AX, AX[test, ])$predictions
}
```

```{r, echo = FALSE}
# 1a) Train Causal Forest
causal_forest <- causal_forest(X = X, Y = Y, W = A,
  Y.hat = Y_X_hat,
  W.hat = A_hat,
  num.trees = 2000,
  min.node.size = 5,
  seed = 123
)
```

```{r, echo = FALSE}
# Extract CATE and variances
tau.hat.oob <- predict(causal_forest, estimate.variance = TRUE)
pension$CATE <- tau.hat.oob$predictions
pension$CATE_se <- sqrt(tau.hat.oob$variance.estimates)

# 95% betrouwbaarheidsintervallen
pension$CATE_lower <- pension$CATE - 1.96 * pension$CATE_se
pension$CATE_upper <- pension$CATE + 1.96 * pension$CATE_se
```

```{r, echo = FALSE}
# plot 95% CI's
ggplot(pension, aes(x = inc, y = CATE)) +
  geom_smooth(aes(ymin = CATE_lower, ymax = CATE_upper), stat = "smooth") +
  labs(title = "Treatment effect vs income (with uncertainty)")
```

```{r, echo = FALSE}
# plot 95% CI's
ggplot(pension, aes(x = age, y = CATE)) +
  geom_smooth(aes(ymin = CATE_lower, ymax = CATE_upper), stat = "smooth") +
  labs(title = "Treatment effect vs income (with uncertainty)")
```


```{r, echo = FALSE}
hist(pension$CATE, main = "CATE")
hist(Y_hat, main = "Expected responses")
hist(A_hat, main = "Propensity scores")
```


```{r, echo = FALSE}
saveRDS(causal_forest, file = "causal_forest_model_ex1_2.rds")
```

```{r, echo = FALSE}
causal_forest2 <- readRDS("causal_forest_model_ex1.rds")
```

```{r, echo = FALSE}
ggplot(pension, aes(x = inc, y = CATE)) +
  geom_point(alpha = 0.2) +
  geom_smooth() +
  labs(title = "Treatment effect vs income")
```

```{r, echo = FALSE}
ggplot(pension, aes(x = age, y = CATE)) +
  geom_point(alpha = 0.2) +
  geom_smooth() +
  labs(title = "Treatment effect vs age")
```

Determining the ATE and his confidence interval.
```{r, echo = FALSE}
ate <- average_treatment_effect(causal_forest,
                               target.sample = "all",
                               method = "AIPW")

ate
```

```{r, echo = FALSE}
ci_low  <- ate[1] - 1.96 * ate[2]
ci_high <- ate[1] + 1.96 * ate[2]
```



## Non-collapsibility in Non-linear Models

First, it is important to note that adding variables to non-linear models generally leads to changes in coefficients, even in the absence of confounding. This phenomenon is known as *non-collapsibility*: marginal and conditional effect measures differ even in the absence of confounding.

## Role of Additional Variables in a Causal Diagram

In addition, including variables in an analysis can have different effects depending on their position in the causal diagram.

- *Collider*  
  In the case of a collider *(Income → Collider ← Outcome)*, conditioning on this variable introduces collider bias.  
  Conditioning on a collider opens a backdoor pathway, making the estimated effect of income biased and very difficult to interpret.

- *Mediator*  
  In the case of a mediator *(Income → Mediator → Outcome)*, part of the effect of income is captured through an indirect effect (via the mediator), and part through a direct effect (the effect observed directly).  
  Including a mediator is therefore not incorrect, but it implies that the estimated effect is no longer the total effect, but only the direct effect.

- *Common Cause*  
  In the case of a common cause *(Income ← Common Cause → Outcome)*, it is appropriate to include this variable in the analysis. If a common cause is omitted, a backdoor pathway remains open, making the estimated effect of income not credible. We note that conditional exchangeability given only income does not hold if there exist common causes of income and the outcome that are not controlled for.

- *Variable Affecting Only Income*  
  If a variable only affects income, including it in the analysis does not introduce bias.  
  However, it may still influence the estimated effect of income due to non-collapsibility in non-linear models and influence the precision of the estimation.

- *Variable Affecting Only the Outcome*  
  If a variable only affects the outcome and not income, it can be included in the analysis.  
  Including such variables may improve the efficiency of the estimation.

## Examples

- *Age* could potentially be a *common cause*.  
  Older individuals tend to earn more and may also be more likely to participate in a 401(k) program.

- *Family Size* could act as a *mediator*, as it may influence the likelihood of participating in a 401(k) program while also being affected by income.

- *Years of Education* could be a *common cause*.  
  The more education an individual has, the higher their income tends to be, and the more likely they may be to participate in a 401(k) program.

## Collider DAG example
\begin{center}
\textbf{Collider} \\
\vspace{0.5cm}
\begin{tikzpicture}[>=stealth, node distance=2cm, thick]
    % Nodes
    \node (A) at (0,2) {Eligibility (A)};
    \node (Y) at (6,2) {Net Assets (Y)};
    \node (Z) at (3,0) { (Collider)};

    % Edges
    \draw[->] (A) -- (Y);
    \draw[->] (A) -- (Z);
    \draw[->] (Y) -- (Z);
\end{tikzpicture}
\end{center}

```{r, echo = FALSE}
#Q2
pension$quartiles <- ntile(pension$CATE, 4)

# Just run the function explicitly for each quartile
Q1_ATE <- average_treatment_effect(causal_forest, subset = (pension$quartiles == 1))
Q2_ATE<- average_treatment_effect(causal_forest, subset = (pension$quartiles == 2))
Q3_ATE <- average_treatment_effect(causal_forest, subset = (pension$quartiles == 3))
Q4_ATE <- average_treatment_effect(causal_forest, subset = (pension$quartiles == 4))


coef_1  <- Q1_ATE["estimate"]
sd_1   <- Q1_ATE["std.err"]

Low_CI_1  <- coef_1 - (1.96 * sd_1)
High_CI_1 <- coef_1 + (1.96 * sd_1)

coef_2  <- Q2_ATE["estimate"]
sd_2   <- Q2_ATE["std.err"]

Low_CI_2  <- coef_2 - (1.96 * sd_2)
High_CI_2 <- coef_2 + (1.96 * sd_2)

coef_3  <- Q3_ATE["estimate"]
sd_3   <- Q3_ATE["std.err"]

Low_CI_3  <- coef_3 - (1.96 * sd_3)
High_CI_3 <- coef_3 + (1.96 * sd_3)

coef_4  <- Q4_ATE["estimate"]
sd_4   <- Q4_ATE["std.err"]

Low_CI_4  <- coef_4 - (1.96 * sd_4)
High_CI_4 <- coef_4 + (1.96 * sd_4)

coef_4
Low_CI_4
High_CI_4

```

```{r, echo = FALSE}
#Q3

## CATE predictions (E(Y1-Y0)/L)
pension$CATE

## prop score (P(A=1/L))
propscore <- A_hat
# zonder cross-fitting: prop_score<-causal_forest$W.hat

print(causal_forest)

## E(Y/L)
var <- Y_X_hat
# zonder cross-fitting: var<-causal_forest$Y.hat


#Get actual average treatment effect (E(Y1-Y0))
average_treatment_effect<-average_treatment_effect(causal_forest)["estimate"]
################################

## E(Y/A,L)
Y_AX_hat

###################################
#dan is het gewoon invullen in die lange EIF formule en kom je er dus wel


#Dit is die eerste term uit de fmle
term1<-(pension$CATE-average_treatment_effect)^2

#weight term, die meest rechtse term uit de fmle
term2<-(W/prop_score)- ((1-W)/(1-prop_score))

#Laatste term uit de fmle, residual
term3<-(Y-xx)
  
#combine
  
formula<-term1+2*(pension$CATE-average_treatment_effect)*term2*term3
#estimate sigma^2
sigma_sq_est<-mean(formula)

#get standard error
ss<-length(formula)
sigma_sq_se<- sd(formula)/sqrt(ss)
#get CI
high_ci<sigma_sq_est-+1.96*sigma_sq_se
low_ci<-sigma_sq_est-1.96*sigma_sq_se





```

```{r, echo = FALSE}
#Q4
# DR-learner implementation
# Reference: solutions of lab Causal Prediction
# wel nog goed verder te bekijken...

Y1_hat = rep(NA, n)
Y0_hat = rep(NA, n)

for (k in 1:K) {

  train <- folds != k
  test <- folds == k
  
  # E[Y|X, A=1]
  rf_Y1 <- regression_forest(X[train & W==1, ], Y[train & W==1, ])
  Y1_hat[test] <- predict(rf_Y1, X[test, ])$predictions
  
  # E[Y|X, A=0]
  rf_Y0 <- regression_forest(X[train & W==0, ], Y[train & W==0, ])
  Y0_hat[test] <- predict(rf_Y0, X[test, ])$predictions
  
  # E[A|X]
  rf_A <- regression_forest(X[train, ], W[train])
  A_hat[test] <- predict(rf_A, X[test, ])$predictions
}
```


```{r, echo = FALSE}
# crossfitting nodig?
###### Compute the pseudooutcomes in the training sets.
C =  #formule cursus

```



```{r, echo = FALSE}
###### Regress the pseudooutcomes in the combined training data set

# crossfitting nodig?
sl_lib = c("SL.glm", "SL.gam", "SL.ranger", "SL.gbm")

sl_fit <- SuperLearner(
  Y = C,
  X = X$inc, # hier enkel de covariate?
  family = gaussian(),
  SL.library = sl_lib,
  method = "method.NNLS"
)

data.01cc.test$CATE.DR <- predict(sl_fit, newdata = x.test, onlySL = TRUE)$pred

####### Plot the treatment effects obtained on the test set on a histogram.
hist(data.01cc.test$CATE.DR, xlab="DR-learner", main="")

```


