---
title: "Causal Machine Learning assignment"
author: "Thomas Sertijn, Bart Smets, Ilja Van Bever"
date: "`r Sys.Date()`"
output: pdf_document
header-includes:
  - \usepackage{tikz}
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(hdm)
library(grf)
library(dplyr)
library(tidyr)
library(knitr)
#load data
data(pension)
library(ggdag)
library(ggplot2)

```

```{r, echo = FALSE}
# Cross-fitting

# Reference doc:
# https://grf-labs.github.io/grf/reference/causal_forest.html
# https://grf-labs.github.io/grf/reference/regression_forest.html

set.seed(123)

variables <- c("net_tfa", "e401", "age", "inc", "fsize", "educ", 
          "marr", "twoearn", "db", "pira", "hown")


Y <- pension$net_tfa
A <- pension$e401
X <- pension[, c("age", "inc", "fsize", "educ", 
                                 "marr", "twoearn", "db", "pira", "hown")]
AX <- pension[, c("e401", "age", "inc", "fsize", "educ", 
                                 "marr", "twoearn", "db", "pira", "hown")]

K <- 5 # set to 5!
n <- nrow(X)
folds <- sample(rep(1:K, length.out = n))

Y_X_hat <- rep(NA, n)
Y_AX_hat <- rep(NA, n)
A_hat <- rep(NA, n)

for (k in 1:K) {
  
  train <- which(folds != k)
  test  <- which(folds == k)
  
  # E[Y|X]
  rf_Y_X <- regression_forest(X[train, ], Y[train])
  Y_X_hat[test] <- predict(rf_Y_X, X[test, ])$predictions
  
  # E[A|X]
  rf_A <- regression_forest(X[train, ], A[train]) # or use probability/classification forest?
  A_hat[test] <- predict(rf_A, X[test, ])$predictions
  # truncation propensity scores toevoegen
  
  # E[Y|A, L] (exercise 3)
  rf_Y_AX <- regression_forest(AX[train, ], Y[train])
  Y_AX_hat[test] <- predict(rf_Y_AX, AX[test, ])$predictions
}
```

```{r, echo = FALSE}
# 1a) Train Causal Forest
causal_forest <- causal_forest(X = X, Y = Y, W = A,
  Y.hat = Y_X_hat,
  W.hat = A_hat,
  num.trees = 2000,
  min.node.size = 5,
  seed = 123
)
```

```{r, echo = FALSE}
# Extract CATE and variances
tau.hat.oob <- predict(causal_forest, estimate.variance = TRUE)
pension$CATE <- tau.hat.oob$predictions
pension$CATE_se <- sqrt(tau.hat.oob$variance.estimates)

# 95% betrouwbaarheidsintervallen
pension$CATE_lower <- pension$CATE - 1.96 * pension$CATE_se
pension$CATE_upper <- pension$CATE + 1.96 * pension$CATE_se
```

```{r, echo = FALSE}
# plot 95% CI's
ggplot(pension, aes(x = inc, y = CATE)) +
  geom_smooth(aes(ymin = CATE_lower, ymax = CATE_upper), stat = "smooth") +
  labs(title = "Treatment effect vs income (with uncertainty)")
```


```{r, echo = FALSE}
# plot 95% CI's
ggplot(pension, aes(x = age, y = CATE)) +
  geom_smooth(aes(ymin = CATE_lower, ymax = CATE_upper), stat = "smooth") +
  labs(title = "Treatment effect vs income (with uncertainty)")
```


```{r, echo = FALSE}
hist(pension$CATE, main = "CATE")
hist(Y_X_hat, main = "Expected responses")
hist(A_hat, main = "Propensity scores")
```
```{r, echo = FALSE}
hist(A_hat, main = "Propensity scores")
```

```{r, echo = FALSE}
saveRDS(causal_forest, file = "causal_forest_model_ex1_2.rds")
```

```{r, echo = FALSE}
causal_forest2 <- readRDS("causal_forest_model_ex1.rds")
```

```{r, echo = FALSE}
ggplot(pension, aes(x = inc, y = CATE)) +
  geom_point(alpha = 0.2) +
  geom_smooth() +
  labs(title = "Treatment effect vs income")
```

```{r, echo = FALSE}
ggplot(pension, aes(x = age, y = CATE)) +
  geom_point(alpha = 0.2) +
  geom_smooth() +
  labs(title = "Treatment effect vs age")
```

Determining the ATE and his confidence interval.
```{r, echo = FALSE}
ate <- average_treatment_effect(causal_forest,
                               target.sample = "all",
                               method = "AIPW")

ate
```

```{r, echo = FALSE}
ci_low  <- ate[1] - 1.96 * ate[2]
ci_high <- ate[1] + 1.96 * ate[2]
```



## Non-collapsibility in Non-linear Models

First, it is important to note that adding variables to non-linear models generally leads to changes in coefficients, even in the absence of confounding. This phenomenon is known as *non-collapsibility*: marginal and conditional effect measures differ even in the absence of confounding.

## Role of Additional Variables in a Causal Diagram

In addition, including variables in an analysis can have different effects depending on their position in the causal diagram.

- *Collider*  
  In the case of a collider *(Income → Collider ← Outcome)*, conditioning on this variable introduces collider bias.  
  Conditioning on a collider opens a backdoor pathway, making the estimated effect of income biased and very difficult to interpret.

- *Mediator*  
  In the case of a mediator *(Income → Mediator → Outcome)*, part of the effect of income is captured through an indirect effect (via the mediator), and part through a direct effect (the effect observed directly).  
  Including a mediator is therefore not incorrect, but it implies that the estimated effect is no longer the total effect, but only the direct effect.

- *Common Cause*  
  In the case of a common cause *(Income ← Common Cause → Outcome)*, it is appropriate to include this variable in the analysis. If a common cause is omitted, a backdoor pathway remains open, making the estimated effect of income not credible. We note that conditional exchangeability given only income does not hold if there exist common causes of income and the outcome that are not controlled for.

- *Variable Affecting Only Income*  
  If a variable only affects income, including it in the analysis does not introduce bias.  
  However, it may still influence the estimated effect of income due to non-collapsibility in non-linear models and influence the precision of the estimation.

- *Variable Affecting Only the Outcome*  
  If a variable only affects the outcome and not income, it can be included in the analysis.  
  Including such variables may improve the efficiency of the estimation.

## Examples

- *Age* could potentially be a *common cause*.  
  Older individuals tend to earn more and may also be more likely to participate in a 401(k) program.

- *Family Size* could act as a *mediator*, as it may influence the likelihood of participating in a 401(k) program while also being affected by income.

- *Years of Education* could be a *common cause*.  
  The more education an individual has, the higher their income tends to be, and the more likely they may be to participate in a 401(k) program.
  
## Collider DAG example
\begin{center}
\textbf{Collider} \\
\vspace{0.5cm}
\begin{tikzpicture}[>=stealth, node distance=2cm, thick]
    % Nodes
    \node (A) at (0,2) {Eligibility (A)};
    \node (Y) at (6,2) {Net Assets (Y)};
    \node (Z) at (3,0) { (Collider)};

    % Edges
    \draw[->] (A) -- (Y);
    \draw[->] (A) -- (Z);
    \draw[->] (Y) -- (Z);
\end{tikzpicture}
\end{center}

```{r, echo = FALSE}

#Q2
pension$quartiles <- ntile(pension$CATE, 4)

# Just run the function explicitly for each quartile
Q1_ATE <- average_treatment_effect(causal_forest, subset = (pension$quartiles == 1))
Q2_ATE <- average_treatment_effect(causal_forest, subset = (pension$quartiles == 2))
Q3_ATE <- average_treatment_effect(causal_forest, subset = (pension$quartiles == 3))
Q4_ATE <- average_treatment_effect(causal_forest, subset = (pension$quartiles == 4))


coef_1  <- Q1_ATE["estimate"]
sd_1   <- Q1_ATE["std.err"]

Low_CI_1  <- coef_1 - (1.96 * sd_1)
High_CI_1 <- coef_1 + (1.96 * sd_1)

coef_2  <- Q2_ATE["estimate"]
sd_2   <- Q2_ATE["std.err"]

Low_CI_2  <- coef_2 - (1.96 * sd_2)
High_CI_2 <- coef_2 + (1.96 * sd_2)

coef_3  <- Q3_ATE["estimate"]
sd_3   <- Q3_ATE["std.err"]

Low_CI_3  <- coef_3 - (1.96 * sd_3)
High_CI_3 <- coef_3 + (1.96 * sd_3)

coef_4  <- Q4_ATE["estimate"]
sd_4   <- Q4_ATE["std.err"]

Low_CI_4  <- coef_4 - (1.96 * sd_4)
High_CI_4 <- coef_4 + (1.96 * sd_4)

coef_4
Low_CI_4
High_CI_4

```

First, we can see that the treatment effect in all four quartiles is positive. This suggests that, on average, 401(k) eligibility has a positive impact on net financial assets across all four quartiles. 

In terms of treatment heterogeneity, we find that the threatment effects for the bottom three quartiles are relatively similar, ranging from $5,826 to $6,437. However, for the 25% of individuals predicted to benefit the most, the treatment effect almost doubles. 

Thus, although the confidence intervals still slightly overlap, still a non-linear treatment effect is suggested. The causal forest did thus succesfully find a sub-population (likely those with higher incomes or higher saving preferences), for which the tax-advantaged saving mechanism is effective.

```{r, echo = FALSE}
#Q3

## CATE predictions (E(Y1-Y0)/L)
CATEhat<-pension$CATE

## prop score (P(A=1/L))
#propscore <- A_hat
propscore <- pmax(0.01, pmin(0.99, A_hat))
# zonder cross-fitting: prop_score<-causal_forest$W.hat

print(causal_forest)

## E(Y/L)
var <- Y_X_hat
# zonder cross-fitting: var<-causal_forest$Y.hat

#Get actual average treatment effect (E(Y1-Y0))
atehat<-average_treatment_effect(causal_forest)["estimate"]
#Get actual average treatment effect (E(Y1-Y0))
atehat <- as.numeric(average_treatment_effect(causal_forest)["estimate"])
## E(Y/A,L)
Y_AX_hat

# Fill in the formula for the EIF

# (Squared difference CATE and ATE) (Term 1)
term1<-(CATEhat-atehat)^2

# Inverse Probability Weights (Term 2)
term2<-(A/propscore)- ((1-A)/(1-propscore))

# Residuals (Term 3)
term3<-(Y-Y_AX_hat)
  
# Combine to get efficient influence function
EIF<-term1+2*(CATEhat-atehat)*term2*term3
# estimate sigma^2, mean(EIF)=0 per definition
sigma_sq_est<-mean(EIF)

#get standard error of sigma_sq_est
# ss = sample size
ss<-length(EIF)
sigma_sq_se<- sd(EIF)/sqrt(ss)
#get CI
high_ci<-sigma_sq_est+1.96*sigma_sq_se
low_ci<-sigma_sq_est-1.96*sigma_sq_se


cat("Estimated Variance (Sigma^2):", sigma_sq_est, "\n")
cat("95% CI: [", low_ci, ",", high_ci, "]\n")

```
The 95% confidence interval still overlaps with 0, so there is no evidence of effect heterogeneity.
```{r, echo = FALSE}
#Q4
# DR-learner implementation
# Assumptions
# 1) heterogeneity only in income (final regression step)
# 2) Doubly robust: als 1 van de nuisance paremeters goed geschat is, is het goed

# Reference: solutions of lab Causal Prediction
# wel nog goed verder te bekijken...

K <- 6 # set to 6 -> 6-fold CF for outer loop and 5-fold DF for inner loop
folds <- sample(rep(1:K, length.out = n))
eps <- 0.01 # epsilon for truncation of propensity-scores

Y1_hat <- matrix(NA, nrow = n, ncol = K)
Y0_hat <- matrix(NA, nrow = n, ncol = K)
A_hat <- matrix(NA, nrow = n, ncol = K)
C_DR <- matrix(NA, nrow = n, ncol = K)

for (l in 1:K) {
  
  k_range <- 1:K
  k_range <- k_range[k_range != l]
  
  for (k in k_range) {
    train_k <- folds %in% k_range & folds != k
    test_k <- folds == k
    
    # E[Y|X, A=1]
    rf_Y1 <- regression_forest(X[train_k & A==1, ], Y[train_k & A==1, ])
    Y1_hat[test_k, l] <- predict(rf_Y1, X[test_k, ])$predictions
    
    # E[Y|X, A=0]
    rf_Y0 <- regression_forest(X[train_k & A==0, ], Y[train_k & A==0, ])
    Y0_hat[test_k, l] <- predict(rf_Y0, X[test_k, ])$predictions
    
    # E[A|X]
    rf_A <- regression_forest(X[train_k, ], A[train_k])
    A_hat_temp <- predict(rf_A, X[test_k, ])$predictions
    A_hat[test_k, l] <- pmin(pmax(A_hat_temp, eps), 1 - eps)
    
    # pseudooutcomes C
    C_DR[test_k, l] <- A[test_k]/A_hat[test_k, l] * (Y[test_k] - Y1_hat[test_k, l]) - (1-A[test_k])/(1-A_hat[test_k, l]) * (Y[test_k] - Y0_hat[test_k, l]) + Y1_hat[test_k, l] - Y0_hat[test_k, l]
    
  }
  
  train <- folds != l
  test <- folds == l
  
  # Regress the pseudooutcomes in the combined training data set
  lm_fit <- lm(C_DR[train, l] ~ X[train, "inc"])

  # Predict CATE on the test set
  CATE[test] <- predict(lm_fit, newdata = data.frame(X[test, "inc"]))
  
}

# ATE
EIF <- rowSums(C_DR)/(K-1)
ate_hat <- mean(EIF)

# The ATE should be equal to the mean of the CATE
mean(CATE)
se_hat  <- sd(EIF) / sqrt(n)
```

ATE = average CATE klopt niet
Betrouwbaarheid ATE obv sd CATES mag niet
Betrouwbaarheidsintervallen CATEs?

```{r, echo = FALSE}
hist(CATE, main = "CATE")
hist(rowmeans(Y1_hat), main = "Expected responses with treatment")
hist(rowmeans(Y0_hat), main = "Expected responses without treatment")
hist(rowmeans(A_hat), main = "Propensity scores")
```

