
---
title: "Causal Machine Learning assignment"
author: "Thomas Sertijn, Bart Smets, Ilja Van Bever"
date: "`r Sys.Date()`"
output: 
pdf_document:
  latex_engine: xelatex
header-includes:
  - \usepackage{tikz}
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(hdm)
library(grf)
library(dplyr)
library(tidyr)
library(knitr)
#load data
data(pension)
library(ggdag)
library(ggplot2)
```

```{r, echo = FALSE}
# Cross-fitting
set.seed(123456789)
# Reference doc:
# https://grf-labs.github.io/grf/reference/causal_forest.html
# https://grf-labs.github.io/grf/reference/regression_forest.html

variables <- c("net_tfa", "e401", "age", "inc", "fsize", "educ", 
          "marr", "twoearn", "db", "pira", "hown")


Y <- pension$net_tfa
A <- pension$e401
X <- pension[, c("age", "inc", "fsize", "educ", 
                                 "marr", "twoearn", "db", "pira", "hown")]
AX <- pension[, c("e401", "age", "inc", "fsize", "educ", 
                                 "marr", "twoearn", "db", "pira", "hown")]

K <- 5 # set to 5!
n <- nrow(X)
folds <- sample(rep(1:K, length.out = n))
# eps <- 0.05 # epsilon for truncation of propensity-scores

Y_X_hat <- rep(NA, n)
Y_AX_hat <- rep(NA, n)
A_hat <- rep(NA, n)

for (k in 1:K) {
  
  train <- which(folds != k)
  test  <- which(folds == k)
  
  # E[Y|X] met regression forests
  rf_Y_X <- regression_forest(X[train, ], Y[train])
  Y_X_hat[test] <- predict(rf_Y_X, X[test, ])$predictions
  
  # E[A|X]
  rf_A <- regression_forest(X[train, ], A[train])
  A_hat[test] <- predict(rf_A, X[test, ])$predictions
  # truncation propensity scores
  #A_hat[test] <- pmin(pmax(A_hat_temp, eps), 1 - eps)
  #A_hat[test] <- A_hat_temp
  
  # E[Y|A, L] (exercise 3)
  rf_Y_AX <- regression_forest(AX[train, ], Y[train])
  Y_AX_hat[test] <- predict(rf_Y_AX, AX[test, ])$predictions
}
```



```{r, echo = FALSE}
set.seed(123456789)
pension$A_hat <- A_hat

ggplot(pension, aes(x = inc, y = A_hat)) +
  geom_point(alpha = 0.2) +
  geom_smooth() +
  labs(title = "Propensity score vs income")
```





Zonder truncation of propensity scores
mean(A_hat < 0.01 | A_hat > 0.99)
[1] 0
mean(A_hat < 0.05 | A_hat > 0.95)
[1] 0.02158346

ATE - Causal Forest = 7892.037
ATE_sd: 1106.29

ATE_CI_low = 5723.7
ATE_CI_high = 10060.38


```{r, echo = FALSE}
set.seed(123456789)
# 1a) Train Causal Forest
causal_forest <- causal_forest(X = X, Y = Y, W = A,
  Y.hat = Y_X_hat,
  W.hat = A_hat,
  num.trees = 2000,
  min.node.size = 5
)
```

```{r, echo = FALSE}
set.seed(123456789)
# Extract CATE and variances
tau.hat.oob <- predict(causal_forest, estimate.variance = TRUE)
pension$CATE <- tau.hat.oob$predictions
```

Hieronder CATE benaderen met smoothing spline (lambda=1)
```{r, echo = FALSE}
set.seed(123456789)
# Smoother
# spline fit
fit <- smooth.spline(x = pension$inc,
                     y = pension$CATE,
                     lambda = 1)

# voorspellingen op een rooster van x-waarden
x_grid <- seq(min(pension$inc),
              max(pension$inc),
              length.out = 500)

pred <- predict(fit, x = x_grid)

```

```{r fig-cate-inc-smooth, echo=FALSE, fig.cap="Treatment effect vs income (smoothing spline)"}
set.seed(123456789)
# plot
# dataframe maken voor ggplot
pred_df <- data.frame(
  inc = pred$x,
  CATE = pred$y
)

# plot
ggplot(pension, aes(x = inc, y = CATE)) +
  geom_point(alpha = 0.2) +
  geom_line(data = pred_df,
            aes(x = inc, y = CATE),
            color = "blue",
            linewidth = 1.2) +
  labs(title = "Treatment effect vs income")
```
En gewoon met lineaire regressie.
```{r fig-cate-inc-lm, echo=FALSE, fig.cap="Treatment effect vs income (linear model)"}
set.seed(123456789)
ggplot(pension, aes(x = inc, y = CATE)) +
  geom_point(alpha = 0.2) +
  geom_smooth(method = "lm",
              se = FALSE,
              color = "blue",
              linewidth = 1.2) +
  labs(title = "Treatment effect vs income")
```

```{r, echo = FALSE}
set.seed(123456789)
pension$CATE_se <- sqrt(tau.hat.oob$variance.estimates)

# 95% betrouwbaarheidsintervallen
pension$CATE_lower <- pension$CATE - 1.96 * pension$CATE_se
pension$CATE_upper <- pension$CATE + 1.96 * pension$CATE_se
```

```{r, echo = FALSE}
set.seed(123456789)
# plot 95% CI's
ggplot(pension, aes(x = inc, y = CATE)) +
  geom_point(alpha = 0.2) +
  geom_smooth(aes(ymin = CATE_lower, ymax = CATE_upper), stat = "smooth") +
  labs(title = "Treatment effect vs income (with uncertainty)")
```

```{r, echo = FALSE}
set.seed(123456789)
# plot 95% CI's
ggplot(pension, aes(x = inc, y = CATE)) +
  geom_smooth(aes(ymin = CATE_lower, ymax = CATE_upper), stat = "smooth") +
  labs(title = "Treatment effect vs income (with uncertainty)")
```

```{r fig-cate-hist, echo=FALSE, fig.cap="Histogram of the predicted CATE values"}
set.seed(123456789)
hist(pension$CATE, main = "CATE", xlab = "CATE")
```

```{r, echo = FALSE}
set.seed(123456789)
hist(Y_X_hat, main = "Expected responses", xlab = "Y")
hist(A_hat, main = "Propensity scores", xlab = "Propensity scores")
```
```{r, echo = FALSE}
set.seed(123456789)
hist(A_hat, main = "Propensity scores")
A_hat_raw <- A_hat  
delta <- 0.05
A_hat[test] <- pmax(delta, pmin(1 - delta, A_hat[test]))

par(mfrow = c(1, 2))
hist(A_hat_raw, breaks = 50, main = "Raw propensity scores",
     xlab = "", col = "lightgrey")
abline(v = c(0.05, 0.95), col = "red", lty = 2)

hist(A_hat, breaks = 50, main = "Truncated propensity scores",
     xlab = "", col = "lightblue")
abline(v = c(0.05, 0.95), col = "red", lty = 2)
par(mfrow = c(1, 1))

cat("Proportion truncated:", mean(A_hat_raw < 0.05 | A_hat_raw > 0.95), "\n")
#Truncation geen groot probleem dus

hist(Y_X_hat, main = "Expected responses", xlab = "Y")
hist(A_hat, main = "Propensity scores", xlab = "Propensity scores")

```

```{r, echo = FALSE}
saveRDS(causal_forest, file = "causal_forest_model_ex1_142_20260509final.rds")
```

```{r, echo = FALSE}
set.seed(123456789)
causal_forest2 <- readRDS("causal_forest_model_ex1.rds")
```

```{r, echo = FALSE}
set.seed(123456789)
ggplot(pension, aes(x = inc, y = CATE)) +
  geom_point(alpha = 0.2) +
  geom_smooth() +
  labs(title = "Treatment effect vs income")
```

```{r, echo = FALSE}
set.seed(123456789)
ggplot(pension, aes(x = age, y = CATE)) +
  geom_point(alpha = 0.2) +
  geom_smooth() +
  labs(title = "Treatment effect vs age")
```

Determining the ATE and his confidence interval.
```{r, echo = FALSE}
set.seed(123456789)
ate <- average_treatment_effect(causal_forest,
                               target.sample = "all",
                               method = "AIPW")

ate
```

```{r, echo = FALSE}
set.seed(123456789)
ci_low  <- ate[1] - 1.96 * ate[2]
ci_high <- ate[1] + 1.96 * ate[2]
```



#1a) interpretation

The causal forest analysis reveals a substantial and statistically significant effect of eligibility for enrollment in a 401(k) plan on accumulated net financial assets. The estimated Average Treatment Effect (ATE) equals 7,892.037, indicating that individuals who are eligible for a 401(k) plan accumulate, on average, approximately \$7,892.037 more in net financial assets than non-eligible individuals. The estimated standard error is \$1,106.29, resulting in a 95% confidence interval of [\$5,723.7, \$10,060.38].

In addition to the average treatment effect, the Conditional Average Treatment Effect (CATE) was examined to investigate heterogeneity in treatment effects across individuals. Figure \@ref(fig:fig-cate-hist) presents a histogram of the predicted CATE values. Most individuals have positive CATE values, although negative values are also observed. Furthermore, the estimated CATE was plotted against income, revealing a clear upward trend: the effect of 401(k) eligibility appears to increase with income. Thus, becoming eligible does have a positive impact on financial assets, but mainly so for richer individuals.

To better characterize this relationship, both a linear regression model and a smoothing spline regression (with $\lambda$ = 1) were fitted. The corresponding trends are displayed in Figures \@ref(fig:fig-cate-inc-smooth) and \@ref(fig:fig-cate-inc-lm) While the estimated curves differ somewhat at higher income levels, this part of the income distribution contains relatively few observations, implying considerably greater uncertainty in this region.

To assess the stability of the inverse probability weights, the distribution of the estimated propensity scores was also investigated using a histogram. No propensity scores below 0.01 or above 0.99 were observed, and only 2.2% of the scores were smaller than 0.05 or larger than 0.95. Overall, the propensity scores do not exhibit problematic extremes. In addition, a clear relationship between income and the propensity scores was observed. This finding is inconsistent with the assumption of exchangeability conditional on income alone. Together with the limited number of extreme propensity scores, this provides an additional reason not to apply propensity score truncation. Removing observations with extreme propensity scores would likely shift the estimated ATE toward the treatment effect for higher-income individuals, thereby reducing the representativeness of the overall estimate. This can be viewed as an example of the bias--variance trade-off. Although trimming extreme propensity scores can reduce the variance of the estimator by limiting the influence of highly weighted observations, it may simultaneously introduce additional bias by systematically excluding specific subpopulations from the analysis.

To estimate the propensity scores and outcomes, 5-fold cross-fitting was applied. The modeling was performed using Regression Forests, as described in the documentation of the GRF package (https://grf-labs.github.io/grf/reference/causal_forest.html).


Nog stukje Bart
Using causal forests with cross-fitting (5 folds), we find the average treatment effect to be approximately: , with a confidence interval ranging from and . Because the confidence interval is positive and larger than 0, we conclude that being eligible for the 401(k) has a significant and positive effect on the net financial assets of a perosn.

In terms of heterogeneity, we find that the treatment effect is not uniform across the population. There is a large, non-linear relationship, as richer individuals have a larger increase in net financial assets when becoming eligible for 401(k) than poorer individuals. Thus, becoming eligible does have a positive impact on financial assets, but mainly so for richer individuals.

#1b)
## Role of Additional Variables in a Causal Diagram

In addition, including variables in an analysis can have different effects depending on their position in the causal diagram.

- *Collider*  
  In the case of a collider *(Eligibility → Collider ← Outcome)*, conditioning on this variable introduces collider bias.  
  Conditioning on a collider opens a backdoor pathway, making the estimated effect of income biased and very difficult to interpret.

- *Mediator*  
  In the case of a mediator *(Eligibility → Mediator → Outcome)*, part of the effect of income is captured through an indirect effect (via the mediator), and part through a direct effect (the effect observed directly).  
  Including a mediator is therefore not incorrect, but it implies that the estimated effect is no longer the total effect, but only the direct effect.

- *Common Cause*  
  In the case of a common cause *(Eligibility ← Common Cause → Outcome)*, it is appropriate to include this variable in the analysis. If a common cause is omitted, a backdoor pathway remains open, making the estimated effect of income not credible. We note that conditional exchangeability given only income does not hold if there exist common causes of income and the outcome that are not controlled for.

- *Variable Affecting Only Income*  
  If a variable only affects income, including it in the analysis does not introduce bias.  
  However, it may still influence the estimated effect of income due to non-collapsibility in non-linear models and influence the precision of the estimation.

- *Variable Affecting Only the Outcome*  
  If a variable only affects the outcome and not income, it can be included in the analysis.  
  Including such variables may improve the efficiency of the estimation.

## Examples

- *Age* could potentially be a *common cause*.  
  Older individuals tend to earn more and may also be more likely to participate in a 401(k) program.

- *Family Size* could act as a *mediator*, as it may influence the likelihood of participating in a 401(k) program while also being affected by income.

- *Years of Education* could be a *common cause*.  
  The more education an individual has, the higher their income tends to be, and the more likely they may be to participate in a 401(k) program.
  
% --- COLLIDER ---
\begin{center}
\textbf{Collider} \\
\vspace{0.5cm}
\begin{tikzpicture}[>=stealth, node distance=2cm, thick]
    \node (A) at (0,2) {Income (A)};
    \node (Y) at (6,2) {Outcome (Y)};
    \node (Z) at (3,0) {Collider (Z)};
    \draw[->] (A) -- (Y);
    \draw[->] (A) -- (Z);
    \draw[->] (Y) -- (Z);
\end{tikzpicture}
\end{center}

% --- MEDIATOR ---
\begin{center}
\textbf{Mediator} \\
\vspace{0.5cm}
\begin{tikzpicture}[>=stealth, node distance=2cm, thick]
    \node (A) at (0,2)  {Income (A)};
    \node (M) at (3,2)  {Mediator (M)};
    \node (Y) at (6,2)  {Outcome (Y)};
    \draw[->] (A) -- (M);
    \draw[->] (M) -- (Y);
    \draw[->, bend right=30] (A) to (Y);  % direct effect
\end{tikzpicture}
\end{center}

% --- COMMON CAUSE ---
\begin{center}
\textbf{Common Cause} \\
\vspace{0.5cm}
\begin{tikzpicture}[>=stealth, node distance=2cm, thick]
    \node (A) at (0,2)  {Income (A)};
    \node (Y) at (6,2)  {Outcome (Y)};
    \node (C) at (3,4)  {Common Cause (C)};
    \draw[->] (A) -- (Y);
    \draw[->] (C) -- (A);
    \draw[->] (C) -- (Y);
\end{tikzpicture}
\end{center}

```{r, echo = FALSE}
set.seed(123456789)
#Q2
pension$quartiles <- ntile(pension$CATE, 4)

# Just run the function explicitly for each quartile
Q1_ATE <- average_treatment_effect(causal_forest, subset = (pension$quartiles == 1))
Q2_ATE <- average_treatment_effect(causal_forest, subset = (pension$quartiles == 2))
Q3_ATE <- average_treatment_effect(causal_forest, subset = (pension$quartiles == 3))
Q4_ATE <- average_treatment_effect(causal_forest, subset = (pension$quartiles == 4))


coef_1  <- Q1_ATE["estimate"]
sd_1   <- Q1_ATE["std.err"]

Low_CI_1  <- coef_1 - (1.96 * sd_1)
High_CI_1 <- coef_1 + (1.96 * sd_1)

coef_2  <- Q2_ATE["estimate"]
sd_2   <- Q2_ATE["std.err"]

Low_CI_2  <- coef_2 - (1.96 * sd_2)
High_CI_2 <- coef_2 + (1.96 * sd_2)

coef_3  <- Q3_ATE["estimate"]
sd_3   <- Q3_ATE["std.err"]

Low_CI_3  <- coef_3 - (1.96 * sd_3)
High_CI_3 <- coef_3 + (1.96 * sd_3)

coef_4  <- Q4_ATE["estimate"]
sd_4   <- Q4_ATE["std.err"]

Low_CI_4  <- coef_4 - (1.96 * sd_4)
High_CI_4 <- coef_4 + (1.96 * sd_4)

coef_1
Low_CI_1
High_CI_1

coef_2
Low_CI_2
High_CI_2

coef_3
Low_CI_3
High_CI_3

coef_4
Low_CI_4
High_CI_4

coef_1
coef_2
coef_3
coef_4

cat("ATE estimates by predicted CATE quartile:\n\n")

cat("Q1 (lowest predicted CATE): ",
    round(coef_1, 2),
    " [", round(Low_CI_1, 2), ", ", round(High_CI_1, 2), "]\n", sep = "")

cat("Q2: ",
    round(coef_2, 2),
    " [", round(Low_CI_2, 2), ", ", round(High_CI_2, 2), "]\n", sep = "")

cat("Q3: ",
    round(coef_3, 2),
    " [", round(Low_CI_3, 2), ", ", round(High_CI_3, 2), "]\n", sep = "")

cat("Q4 (highest predicted CATE): ",
    round(coef_4, 2),
    " [", round(Low_CI_4, 2), ", ", round(High_CI_4, 2), "]\n", sep = "")

```

First, we can see that the treatment effect in all four quartiles is positive. This suggests that, on average, 401(k) eligibility has a positive impact on net financial assets across all four quartiles. 

In terms of treatment heterogeneity, we find that the threatment effects for the bottom three quartiles are relatively similar, ranging from $5,826 to $6,437. However, for the 25% of individuals predicted to benefit the most, the treatment effect almost doubles. 

Thus, although the confidence intervals still slightly overlap, still a non-linear treatment effect is suggested. The causal forest did thus succesfully find a sub-population (likely those with higher incomes or higher saving preferences), for which the tax-advantaged saving mechanism is effective.

```{r, echo = FALSE}
set.seed(123456789)
#Q3

## CATE predictions (E(Y1-Y0)/L)
CATEhat<-pension$CATE

## prop score (P(A=1/L))
propscore <- A_hat # propensity scores are already truncated in 1a)

# zonder cross-fitting: prop_score<-causal_forest$W.hat

print(causal_forest)

## E(Y/L)
var <- Y_X_hat
# zonder cross-fitting: var<-causal_forest$Y.hat

#Get actual average treatment effect (E(Y1-Y0))
atehat<-average_treatment_effect(causal_forest)["estimate"]
#Get actual average treatment effect (E(Y1-Y0))
atehat <- as.numeric(average_treatment_effect(causal_forest)["estimate"])
## E(Y/A,L)
Y_AX_hat

# Fill in the formula for the EIF

# (Squared difference CATE and ATE) (Term 1)
term1<-(CATEhat-atehat)^2

# Inverse Probability Weights (Term 2)
term2<-(A/propscore)- ((1-A)/(1-propscore))

# Residuals (Term 3)
term3<-(Y-Y_AX_hat)
  
# Combine to get efficient influence function
sigma_sq_hat<-mean(term1+2*(CATEhat-atehat)*term2*term3)
# estimate sigma^2, mean(EIF)=0 per definition
IF_sigma<- term1 - sigma_sq_hat +2 * (CATEhat - atehat) * term2 * term3
#get standard error of sigma_sq_est
# ss = sample size
ss<-length(IF_sigma)
sigma_sq_se<- sd(IF_sigma)/sqrt(ss)
#get CI
high_ci<-sigma_sq_hat+1.96*sigma_sq_se
low_ci<-sigma_sq_hat-1.96*sigma_sq_se


cat("Estimated Variance (Sigma^2):", sigma_sq_hat, "\n")
cat("95% CI: [", low_ci, ",", high_ci, "]\n")
```

```{r, echo = FALSE}
# ============================================================
# 1. Inspect largest absolute influence-function values
# ============================================================

correction <- 2 * (CATEhat - atehat) * term2 * term3

diag_IF <- data.frame(
  row_id = seq_along(Y),
  net_tfa = Y,
  A = A,
  CATE = CATEhat,
  CATE_minus_ATE = CATEhat - atehat,
  propensity = propscore,
  ipw_contrast = term2,
  residual = term3,
  correction = correction,
  IF_sigma = IF_sigma,
  abs_IF_sigma = abs(IF_sigma)
)

top_IF <- diag_IF %>%
  arrange(desc(abs_IF_sigma)) %>%
  slice(1:10)

knitr::kable(
  top_IF,
  digits = 2,
  caption = "Ten observations with largest absolute influence-function values"
)

# ============================================================
# 2. Decompose influence-function components
# ============================================================

component_summary <- data.frame(
  Component = c(
    "Squared CATE deviation",
    "Correction term",
    "Influence function"
  ),
  Mean = c(
    mean(term1, na.rm = TRUE),
    mean(correction, na.rm = TRUE),
    mean(IF_sigma, na.rm = TRUE)
  ),
  SD = c(
    sd(term1, na.rm = TRUE),
    sd(correction, na.rm = TRUE),
    sd(IF_sigma, na.rm = TRUE)
  ),
  Min = c(
    min(term1, na.rm = TRUE),
    min(correction, na.rm = TRUE),
    min(IF_sigma, na.rm = TRUE)
  ),
  Max = c(
    max(term1, na.rm = TRUE),
    max(correction, na.rm = TRUE),
    max(IF_sigma, na.rm = TRUE)
  ),
  Q99_abs = c(
    quantile(abs(term1), 0.99, na.rm = TRUE),
    quantile(abs(correction), 0.99, na.rm = TRUE),
    quantile(abs(IF_sigma), 0.99, na.rm = TRUE)
  )
)

knitr::kable(
  component_summary,
  digits = 2,
  caption = "Decomposition of the influence-function estimator"
)

driver_correlations <- data.frame(
  Driver = c(
    "|residual|",
    "|IPW contrast|",
    "|CATE - ATE|",
    "|correction term|"
  ),
  Correlation_with_abs_IF = c(
    cor(abs(term3), abs(IF_sigma), use = "complete.obs"),
    cor(abs(term2), abs(IF_sigma), use = "complete.obs"),
    cor(abs(CATEhat - atehat), abs(IF_sigma), use = "complete.obs"),
    cor(abs(correction), abs(IF_sigma), use = "complete.obs")
  )
)

knitr::kable(
  driver_correlations,
  digits = 3,
  caption = "Correlation between absolute influence-function values and possible drivers"
)

# ============================================================
# 3. Check whether large residuals are driven by high net_tfa values
# ============================================================

diag_IF <- diag_IF %>%
  mutate(
    abs_residual = abs(residual),
    abs_ipw = abs(ipw_contrast),
    top_IF_1pct = abs_IF_sigma >= quantile(abs_IF_sigma, 0.99, na.rm = TRUE),
    top_resid_1pct = abs_residual >= quantile(abs_residual, 0.99, na.rm = TRUE),
    top_y_1pct = net_tfa >= quantile(net_tfa, 0.99, na.rm = TRUE)
  )

outlier_summary <- diag_IF %>%
  group_by(top_IF_1pct) %>%
  summarise(
    n = n(),
    mean_net_tfa = mean(net_tfa, na.rm = TRUE),
    median_net_tfa = median(net_tfa, na.rm = TRUE),
    mean_abs_residual = mean(abs_residual, na.rm = TRUE),
    median_abs_residual = median(abs_residual, na.rm = TRUE),
    mean_abs_ipw = mean(abs_ipw, na.rm = TRUE),
    min_propensity = min(propensity, na.rm = TRUE),
    max_propensity = max(propensity, na.rm = TRUE),
    prop_top_y_1pct = mean(top_y_1pct, na.rm = TRUE),
    prop_top_resid_1pct = mean(top_resid_1pct, na.rm = TRUE),
    .groups = "drop"
  )

knitr::kable(
  outlier_summary,
  digits = 3,
  caption = "Comparison of top 1% IF observations with the rest of the sample"
)

ggplot(diag_IF, aes(x = net_tfa, y = abs_residual)) +
  geom_point(alpha = 0.25, size = 0.7) +
  geom_smooth(se = FALSE) +
  scale_x_continuous(trans = "pseudo_log") +
  scale_y_continuous(trans = "pseudo_log") +
  theme_minimal() +
  labs(
    title = "Large residuals are driven by high net financial assets",
    x = "Net financial assets",
    y = expression("|Y - " * hat(E)(Y ~ "|" ~ A * "," * L) * "|")
  )
compact_explanation <- diag_IF %>%
  group_by(top_IF_1pct) %>%
  summarise(
    n = n(),
    mean_net_tfa = mean(net_tfa),
    mean_abs_residual = mean(abs_residual),
    mean_abs_ipw = mean(abs_ipw),
    min_propensity = min(propensity),
    max_propensity = max(propensity),
    .groups = "drop"
  )

knitr::kable(
  compact_explanation,
  digits = 2,
  caption = "Main driver of extreme influence-function values"
)

# ============================================================
# Decompose the correction term into its three components
# ============================================================

cate_component  <- 2 * (CATEhat - atehat)
ipw_component   <- term2
resid_component <- term3

correction <- cate_component * ipw_component * resid_component

correction_diag <- data.frame(
  row_id = seq_along(Y),
  net_tfa = Y,
  IF_sigma = IF_sigma,
  abs_IF = abs(IF_sigma),
  correction = correction,
  abs_correction = abs(correction),
  cate_component = cate_component,
  abs_cate_component = abs(cate_component),
  ipw_component = ipw_component,
  abs_ipw_component = abs(ipw_component),
  resid_component = resid_component,
  abs_resid_component = abs(resid_component),
  propensity = propscore
)

# Show the most influential observations
top_correction <- correction_diag %>%
  arrange(desc(abs_IF)) %>%
  slice(1:10)

knitr::kable(
  top_correction,
  digits = 2,
  caption = "Decomposition of the correction term for the largest IF values"
)

```

```{r, echo = FALSE}
# # ============================================================
# # Assess drivers of extreme influence-function values
# # ============================================================
# 
# # Components
# cate_dev        <- CATEhat - atehat
# abs_cate_dev    <- abs(cate_dev)
# 
# squared_term    <- cate_dev^2
# 
# ipw_contrast    <- term2
# abs_ipw         <- abs(ipw_contrast)
# 
# residual        <- term3
# abs_residual    <- abs(residual)
# 
# correction_term <- 2 * cate_dev * ipw_contrast * residual
# 
# abs_IF          <- abs(IF_sigma)
# 
# # Diagnostic dataset
# IF_drivers <- data.frame(
#   row_id = seq_along(Y),
#   net_tfa = Y,
#   IF_sigma = IF_sigma,
#   abs_IF = abs_IF,
#   CATE = CATEhat,
#   ATE = atehat,
#   cate_dev = cate_dev,
#   abs_cate_dev = abs_cate_dev,
#   squared_term = squared_term,
#   abs_squared_term = abs(squared_term),
#   ipw_contrast = ipw_contrast,
#   abs_ipw = abs_ipw,
#   residual = residual,
#   abs_residual = abs_residual,
#   correction_term = correction_term,
#   abs_correction_term = abs(correction_term),
#   propensity = propscore
# )
# 
# # Define extreme IF values: top 1%
# cut_IF_99 <- quantile(IF_drivers$abs_IF, 0.99, na.rm = TRUE)
# 
# IF_drivers <- IF_drivers %>%
#   mutate(top_IF_1pct = abs_IF >= cut_IF_99)
# 
# # ============================================================
# # 1. Compare top 1% IF observations with the rest
# # ============================================================
# 
# IF_driver_summary <- IF_drivers %>%
#   group_by(top_IF_1pct) %>%
#   summarise(
#     n = n(),
#     mean_abs_IF = mean(abs_IF, na.rm = TRUE),
#     mean_abs_cate_dev = mean(abs_cate_dev, na.rm = TRUE),
#     mean_squared_term = mean(squared_term, na.rm = TRUE),
#     mean_abs_ipw = mean(abs_ipw, na.rm = TRUE),
#     mean_abs_residual = mean(abs_residual, na.rm = TRUE),
#     mean_abs_correction = mean(abs_correction_term, na.rm = TRUE),
#     mean_net_tfa = mean(net_tfa, na.rm = TRUE),
#     min_propensity = min(propensity, na.rm = TRUE),
#     max_propensity = max(propensity, na.rm = TRUE),
#     .groups = "drop"
#   )
# 
# knitr::kable(
#   IF_driver_summary,
#   digits = 2,
#   caption = "Comparison of top 1% influence-function observations with the rest"
# )
# 
# # ============================================================
# # 2. Correlations with absolute IF values
# # ============================================================
# 
# driver_correlations <- data.frame(
#   Driver = c(
#     "|CATE - ATE|",
#     "Squared CATE-deviation term",
#     "|IPW contrast|",
#     "|Residual|",
#     "|Correction term|"
#   ),
#   Correlation_with_abs_IF = c(
#     cor(IF_drivers$abs_cate_dev, IF_drivers$abs_IF, use = "complete.obs"),
#     cor(IF_drivers$abs_squared_term, IF_drivers$abs_IF, use = "complete.obs"),
#     cor(IF_drivers$abs_ipw, IF_drivers$abs_IF, use = "complete.obs"),
#     cor(IF_drivers$abs_residual, IF_drivers$abs_IF, use = "complete.obs"),
#     cor(IF_drivers$abs_correction_term, IF_drivers$abs_IF, use = "complete.obs")
#   )
# )
# 
# knitr::kable(
#   driver_correlations,
#   digits = 3,
#   caption = "Correlation between absolute IF values and possible drivers"
# )
# 
# # ============================================================
# # 3. Inspect the most influential observations
# # ============================================================
# 
# top_IF_observations <- IF_drivers %>%
#   arrange(desc(abs_IF)) %>%
#   slice(1:10) %>%
#   select(
#     row_id,
#     net_tfa,
#     IF_sigma,
#     abs_IF,
#     cate_dev,
#     squared_term,
#     ipw_contrast,
#     residual,
#     correction_term,
#     propensity
#   )
# 
# knitr::kable(
#   top_IF_observations,
#   digits = 2,
#   caption = "Ten observations with largest absolute influence-function values"
# )

# Compact diagnostic table: what drives extreme IF values?

# Compact report table: drivers of extreme IF values

cate_dev <- CATEhat - atehat
correction_term <- 2 * cate_dev * term2 * term3

IF_drivers <- data.frame(
  abs_IF = abs(IF_sigma),
  abs_CATE_dev = abs(cate_dev),
  abs_IPW = abs(term2),
  abs_residual = abs(term3),
  net_tfa = Y,
  propensity = propscore
)

cut_IF_99 <- quantile(IF_drivers$abs_IF, 0.99, na.rm = TRUE)

IF_drivers <- IF_drivers %>%
  mutate(
    group = ifelse(abs_IF >= cut_IF_99,
                   "Top 1% |IF|",
                   "Remaining 99%")
  )

IF_report_table <- IF_drivers %>%
  group_by(group) %>%
  summarise(
    n = n(),
    `Mean net financial assets` = mean(net_tfa, na.rm = TRUE),
    `Mean |residual|` = mean(abs_residual, na.rm = TRUE),
    `Mean |CATE - ATE|` = mean(abs_CATE_dev, na.rm = TRUE),
    `Mean |IPW contrast|` = mean(abs_IPW, na.rm = TRUE),
    `Propensity range` = paste0(
      round(min(propensity, na.rm = TRUE), 2),
      "–",
      round(max(propensity, na.rm = TRUE), 2)
    ),
    .groups = "drop"
  ) %>%
  arrange(desc(group == "Top 1% |IF|"))

knitr::kable(
  IF_report_table,
  digits = 2,
  caption = "Comparison of the most influential observations with the rest of the sample"
)

```
sigma_sq_hat
16169595
sigma_sq_se
41439971
sigma_sq_hat+1.96*sigma_sq_se
[1] 97391939
sigma_sq_hat-1.96*sigma_sq_se
[1] -65052749


As CATE variance estimate (and its confidence intervals) are very large; (\(\hat{\sigma}^2 = `r round(sigma_sq_hat, 2)`\), 95% CI 
[`r round(low_ci, 2)`, `r round(high_ci, 2)`]), we also inspect what drives this value. For this purpose, we inspect whether this result is driven by a few observations with extreme EIF values, which we define as values in the top percent of efficient influence function values. The largest absolute influence-function value equals 
`r round(max(abs(IF_sigma), na.rm = TRUE), 2)`, while the 99th percentile equals 
`r round(quantile(abs(IF_sigma), 0.99, na.rm = TRUE), 2)`. 

These extreme EIF values can originate from three 'sources'. First, they may come from large deviations of the estimated CATE from the ATE. Second, they can come from extreme propensity scores. Third, they can be the result from large outcome residuals. Hence, we compare these for the top 1% vs top 99% of the population.

As shown in Table X, the extreme EIF values are the result of large residuals. The mean absolute residual among the top 1% most influential observations is 
`r round(IF_report_table$`Mean |residual|`[IF_report_table$group == "Top 1% |IF|"], 2)`, compared with 
`r round(IF_report_table$`Mean |residual|`[IF_report_table$group == "Remaining 99%"], 2)` for the rest of the sample. These extreme values for these observations are the result of higher net financial assets on average: 
`r round(IF_report_table$`Mean net financial assets`[IF_report_table$group == "Top 1% |IF|"], 2)`, compared with 
`r round(IF_report_table$`Mean net financial assets`[IF_report_table$group == "Remaining 99%"], 2)`. This indicates that several high-asset households have observed outcomes that are substantially larger than predicted by the outcome model. On the other hand, the CATE-ATE deviations are also large, but smaller. The propensity scores among these extreme EIF values are rather normal.

Overall, this suggests that the wide confidence interval is not primarily caused by severe overlap violations, but by a small number of high-asset observations with very large outcome residuals. Since the debiased estimator is not constrained to be non-negative in finite samples, these large correction terms can dominate the positive squared CATE-deviation term. Therefore, the evidence for treatment-effect heterogeneity should be interpreted cautiously.




Extreme values for these influence functions can either be driven by the ... term, the propensity scores or the residuals. Therefore, we also inspect whether our extreme influence function values are driven by these. As we can see in table x, these are mainly driven by  very large residuals. This is because certain observations have very large residuals, as their net financial assets is substantially larger than what's predicted. For example,...

```{r, echo = FALSE}
set.seed(123456789)
#Extreme influence function diagnostics
summary(IF_sigma)

# Quantiles 
quantile(abs(IF_sigma), probs = c(0.50, 0.90, 0.95, 0.975, 0.99, 0.995, 1))

# Histogram of influence-function values
hist(IF_sigma,
     breaks = 50,
     main = "Estimated influence function for sigma^2",
     xlab = "Influence-function value")

# Histogram of absolute influence-function values
hist(abs(IF_sigma),
     breaks = 50,
     main = "Absolute influence-function values",
     xlab = "|Influence-function value|")

# most influential observations
abs_IF <- abs(IF_sigma)

top_ids <- order(abs_IF, decreasing = TRUE)[1:20]

top_influential <- pension[top_ids, c(
  "net_tfa", "e401", "age", "inc", "fsize", "educ",
  "marr", "twoearn", "db", "pira", "hown"
)]

top_influential$IF_sigma <- IF_sigma[top_ids]
top_influential$abs_IF_sigma <- abs_IF[top_ids]
top_influential$CATE <- CATEhat[top_ids]
top_influential$ATE <- atehat
top_influential$CATE_minus_ATE <- CATEhat[top_ids] - atehat
top_influential$propensity <- propscore[top_ids]
top_influential$ipw_contrast <- term2[top_ids]
top_influential$residual <- term3[top_ids]

top_influential <- top_influential[order(-top_influential$abs_IF_sigma), ]

knitr::kable(top_influential, digits = 2)
diagnostic_components <- data.frame(
  net_tfa = Y,
  A = A,
  CATE = CATEhat,
  CATE_minus_ATE = CATEhat - atehat,
  propensity = propscore,
  ipw_contrast = term2,
  residual = term3,
  term1 = term1,
  correction = 2 * (CATEhat - atehat) * term2 * term3,
  IF_sigma = IF_sigma,
  abs_IF_sigma = abs(IF_sigma)
)

top_components <- diagnostic_components %>%
  arrange(desc(abs_IF_sigma)) %>%
  slice(1:20)

knitr::kable(top_components, digits = 2)

summary(pension$net_tfa)

quantile(pension$net_tfa,
         probs = c(0, 0.5, 0.9, 0.95, 0.99, 0.995, 1))

hist(pension$net_tfa,
     breaks = 50,
     main = "Distribution of net financial assets",
     xlab = "net_tfa")

```
```{r, echo = FALSE}
# Propensity against covariates
p_hat <- A_hat
par(mfrow = c(1, 2))
plot(X$inc, p_hat, pch = 16, cex = 0.3, col = rgb(0,0,0,0.3),
     xlab = "Income", ylab = expression(hat(e)(L)),
     main = "Propensity vs income")
abline(h = c(0.05, 0.95), col = "red", lty = 2)

plot(X$age, p_hat, pch = 16, cex = 0.3, col = rgb(0,0,0,0.3),
     xlab = "Age", ylab = expression(hat(e)(L)),
     main = "Propensity vs age")
abline(h = c(0.05, 0.95), col = "red", lty = 2)
par(mfrow = c(1, 1))

p_hat_trunc <- pmax(0.05, pmin(0.95, p_hat))

plot_df <- as.data.frame(X)
plot_df$p_hat_raw   <- p_hat
plot_df$p_hat_trunc <- p_hat_trunc

plot_long <- plot_df %>%
  pivot_longer(cols = -c(p_hat_raw, p_hat_trunc),
               names_to = "covariate", values_to = "value")

ggplot(plot_long, aes(x = value, y = p_hat_raw)) +
  geom_point(alpha = 0.1, size = 0.3) +
  geom_hline(yintercept = c(0.05, 0.95), col = "red", lty = 2) +
  facet_wrap(~ covariate, scales = "free_x") +
  labs(title = "Raw propensity vs all covariates",
       y = expression(hat(e)(L)), x = "Covariate value")


ggplot(plot_long, aes(x = value, y = p_hat_trunc)) +
  geom_point(alpha = 0.1, size = 0.3) +
  geom_hline(yintercept = c(0.05, 0.95), col = "red", lty = 2) +
  facet_wrap(~ covariate, scales = "free_x") +
  labs(title = "Truncated propensity vs all covariates",
       y = expression(hat(e)(L)), x = "Covariate value")


ggplot(plot_df, aes(x = inc, y = p_hat_raw)) +
  geom_point(alpha = 0.1, size = 0.3) +
  geom_hline(yintercept = c(0.05, 0.95), col = "red", lty = 2) +
  labs(title = "Propensity vs income (raw vs truncated)",
       y = expression(hat(e)(L)), x = "Income") +
  geom_point(aes(y = p_hat_trunc), alpha = 0.1, size = 0.3, col = "blue")
```
```{r}
# Overlap plot: propensity score distribution by treatment group

pension$p_hat_raw <- A_hat

pension$e401_factor <- factor(
  pension$e401,
  levels = c(0, 1),
  labels = c("Not eligible", "Eligible")
)

ggplot(pension, aes(x = p_hat_raw, fill = e401_factor, color = e401_factor)) +
  geom_density(alpha = 0.3, linewidth = 0.8) +
  geom_vline(xintercept = c(0.05, 0.95), linetype = "dashed", color = "red") +
  theme_minimal() +
  labs(
    title = "Propensity score overlap by treatment group",
    x = expression(hat(P)(A == 1 ~ "|" ~ L)),
    y = "Density",
    fill = "401(k) eligibility",
    color = "401(k) eligibility"
  )
```





(The 95% confidence interval still overlaps with 0, so there is no evidence of effect heterogeneity.)

To assess treatment effect heterogeneity, the variance of the CATE was calculated as given in the assignment. This gives us an idea of the extent as to which the impact of 401(k) eligibility on the net amount of financial assets is different across households with different covariate characteristics. If the effect is close to 0, then we would conclude that the treatment effect is homogenous for households characterized by similar covariates.

More formally, we estimated:

\[
\sigma^2 = Var\{E(Y^1 - Y^0 \mid L)\},
\]

where \(Y^1\) and \(Y^0\) give the outcomes under treatment (eligibility) and control (non-eligibility), respectively, and \(L\) gives the covariates:

\[
\tau(L) = E(Y^1 - Y^0 \mid L),
\]

Thus, we can write:

\[
\sigma^2 = Var\{\tau(L)\}.
\]

If

\[
\psi = E\{\tau(L)\}
\]

gives the average treatment effect. Then,

\[
\sigma^2 = E\left[\{\tau(L)-\psi\}^2\right].
\]

Thus, \(\sigma^2\) measures the variance of the CATE around the ATE. 

Normally, we would just compute the variance around the CATE's. However, CATEs are estimated and therefore, this could lead to incorrect inference. Therefore, the efficient influence function for \(\sigma^2\) is used, denoted by:

\[
\begin{aligned}
\phi(O) 
=& \{\tau(L)-\psi\}^2 - \sigma^2 \\
&+ 2\{\tau(L)-\psi\}
\left\{
\frac{A}{P(A=1\mid L)}
-
\frac{1-A}{P(A=0\mid L)}
\right\}
\{Y - E(Y\mid A,L)\}.
\end{aligned}
\]

The efficient influence function has mean zero at the true value of the parameter, meaning that

\[
E\{\phi(O)\}=0.
\]

This property can be used to construct an estimator for \(\sigma^2\). In practice, we replace the unknown nuisance functions by estimates and solve the empirical analogue of this mean-zero condition:

\[
\frac{1}{n}\sum_{i=1}^{n}\hat{\phi}_i = 0.
\]

Starting from the efficient influence function, this means solving

\[
\frac{1}{n}\sum_{i=1}^{n}
\left[
\{\hat{\tau}(L_i)-\hat{\psi}\}^2 - \hat{\sigma}^2
+
2\{\hat{\tau}(L_i)-\hat{\psi}\}
\left\{
\frac{A_i}{\hat{p}(L_i)}
-
\frac{1-A_i}{1-\hat{p}(L_i)}
\right\}
\{Y_i-\hat{m}(A_i,L_i)\}
\right]
=0.
\]



We replace the (unknown) nuisance functions by empirical estimates

\[
\hat{\tau}(L_i)
\]

for the estimated CATE from the causal forest,

\[
\hat{\psi}
\]

for the estimated average treatment effect,

\[
\hat{p}(L_i) = \hat{P}(A_i=1\mid L_i)
\]

for the estimated propensity score,

\[
\hat{m}(A_i,L_i) = \hat{E}(Y_i\mid A_i,L_i)
\]

for the estimated outcome regression.

This gives the following debiased estimator:

\[
\hat{\sigma}^2 =
\frac{1}{n}\sum_{i=1}^{n}
\left[
\{\hat{\tau}(L_i)-\hat{\psi}\}^2
+
2\{\hat{\tau}(L_i)-\hat{\psi}\}
\left\{
\frac{A_i}{\hat{p}(L_i)}
-
\frac{1-A_i}{1-\hat{p}(L_i)}
\right\}
\{Y_i-\hat{m}(A_i,L_i)\}
\right].
\]

The first term,

\[
\{\hat{\tau}(L_i)-\hat{\psi}\}^2,
\]

gives the estimated variation in the CATE around the ATE. The second term is a debiasing correction based on the residuals from the outcome regression and inverse probability weighting. This correction accounts for potential misspecification in the nuisance functions.

Then, we estimated the influence-function values as

\[
\begin{aligned}
\hat{\phi}_i 
=& \{\hat{\tau}(L_i)-\hat{\psi}\}^2 - \hat{\sigma}^2 \\
&+ 2\{\hat{\tau}(L_i)-\hat{\psi}\}
\left\{
\frac{A_i}{\hat{p}(L_i)}
-
\frac{1-A_i}{1-\hat{p}(L_i)}
\right\}
\{Y_i-\hat{m}(A_i,L_i)\}.
\end{aligned}
\]

The standard error is estimated as:

\[
\widehat{SE}(\hat{\sigma}^2)
=
\frac{sd(\hat{\phi}_i)}{\sqrt{n}},
\]

and a 95% confidence interval was constructed as

\[
\hat{\sigma}^2
\pm
1.96 \times \widehat{SE}(\hat{\sigma}^2).
\]

Ultimately, we inspected the estimated influence-function values and the inverse probability weights as diagnostics, to see if our estimates and CI's were driven by those.

#Main conclusion

Our estimate for the variance is approximately −6.68×10^6 (se=...), with a confidence interval around it of ... and .... As σ^2 denotes the variance, this obviously cannot be negative in reality. However, as we use the debiased estimator, this is not constrained to be negative. To see this, consider that the first term of the EIF should always be positive (as it is squared), however if the correction term is larger than the first term, then the found estimate can be negative. 

Our estimate of ... is thus positive.  

The 95% confidence interval includes 0, and this thus means that we cannot concldue that there is significant heterogeneity in the impact of being eligible for 401(k) schemes for different households.

As our estimates are rather large, we also took a deeper look at the values for the efficient influence functions. This check indicated that certain observations had large absolute influence function values. More specifically, these values are mostly driven by households with a high amount of net financial assets (and thus large residuals). We further investigated whether the propensity scores of these observations were close to 0 or 1, and this was not the case. Thus, these values were caused by the very large residual values caused by the right-skewed distribution, not by the propensity scores.

Overall, as the confidence intervals overlap with 0, we cannot conclude treatment effect heterogeneity. However, as these confidence intervals remain very broad, it is also difficult to conclude that there is no heterogeneity. Therefore, it is difficult to take conclusions, as inference is impacted by specific high-asset households.

Q4 met DR-learner
```{r, echo = FALSE}
set.seed(123456789)
# Q4
# DR-learner implementation
# Assumptions
# 1) heterogeneity only in income (final regression step)
# 2) Doubly robust: als 1 van de nuisance paremeters goed geschat is, is het goed

# Reference: solutions of lab Causal Prediction
n <- nrow(X)
K <- 6 # set to 6 -> 6-fold CF for outer loop and 5-fold DF for inner loop

variables <- c("net_tfa", "e401", "age", "inc", "fsize", "educ", 
          "marr", "twoearn", "db", "pira", "hown")

Y <- pension$net_tfa
A <- pension$e401
X <- pension[, c("age", "inc", "fsize", "educ", 
                                 "marr", "twoearn", "db", "pira", "hown")]
AX <- pension[, c("e401", "age", "inc", "fsize", "educ", 
                               "marr", "twoearn", "db", "pira", "hown")]

eps <- 0.01 # epsilon for truncation of propensity-scores

Y1_hat <- matrix(NA, nrow = n, ncol = K)
Y0_hat <- matrix(NA, nrow = n, ncol = K)
A_hat <- matrix(NA, nrow = n, ncol = K)
C_DR <- matrix(NA, nrow = n, ncol = K)
CATE <- rep(NA, n)
```

Q4 met DR-learner

```{r, echo = FALSE}
set.seed(123456789)
# Q4
# Random folds
# folds <- sample(rep(1:K, length.out = n))

# Stratified folds
folds <- rep(NA, n)

# indices per class
idx0 <- which(A == 0)
idx1 <- which(A == 1)

# random sampling within classes
idx0 <- sample(idx0)
idx1 <- sample(idx1)

# folds
folds[idx0] <- rep(1:K, length.out = length(idx0))
folds[idx1] <- rep(1:K, length.out = length(idx1))
```

```{r, echo = FALSE}
set.seed(123456789)
# Q4
# DR-learner implementation
# Loop

for (l in 1:K) {
  
  k_range <- 1:K
  k_range <- k_range[k_range != l]
  
  for (k in k_range) {
    train_k <- folds %in% k_range & folds != k
    test_k <- folds == k
    
    # E[Y|X, A=1]
    rf_Y1 <- regression_forest(X[train_k & A==1, ], Y[train_k & A==1])
    Y1_hat[test_k, l] <- predict(rf_Y1, X[test_k, ])$predictions
    
    # E[Y|X, A=0]
    rf_Y0 <- regression_forest(X[train_k & A==0, ], Y[train_k & A==0])
    Y0_hat[test_k, l] <- predict(rf_Y0, X[test_k, ])$predictions
    
    # E[A|X]
    rf_A <- regression_forest(X[train_k, ], A[train_k])
    A_hat_temp <- predict(rf_A, X[test_k, ])$predictions
    A_hat[test_k, l] <- pmin(pmax(A_hat_temp, eps), 1 - eps)
    
    # pseudooutcomes C
    C_DR[test_k, l] <- A[test_k]/A_hat[test_k, l] * (Y[test_k] - Y1_hat[test_k, l]) - (1-A[test_k])/(1-A_hat[test_k, l]) * (Y[test_k] - Y0_hat[test_k, l]) + Y1_hat[test_k, l] - Y0_hat[test_k, l]
    
  }
  
  train <- folds != l
  test <- folds == l
  
  # training dataframe
  train_df <- data.frame(
    C_DR = C_DR[train, l],
    inc = X[train, "inc"]
  )
  
  # omit columns with NA values before regression
  # train_df <- na.omit(train_df)#
  
  # Regress the pseudooutcomes in the combined training data set
  lm_fit <- lm(C_DR ~ inc, data = train_df)

  # test dataframe
  test_df <- data.frame(
    inc = X[test, "inc"]
  )

  # Predict CATE on the test set
  CATE[test] <- predict(lm_fit, newdata = test_df)
}


```

```{r, echo = FALSE}
set.seed(123456789)
l <- 1
train <- folds != l
test <- folds == l

# training dataframe
train_df <- data.frame(
  C_DR = C_DR[train, l],
  inc = X[train, "inc"]
)

# omit columns with NA values before regression
# train_df <- na.omit(train_df)

# Plot
ggplot(train_df, aes(x = inc, y = C_DR)) +
geom_point(alpha = 0.2) +
geom_smooth(method = "lm") +
coord_cartesian(ylim = c(0, 60000)) +
labs(title = "Treatment effect vs income")

```
```{r, echo = FALSE}
l <- 2
train <- folds != l
test <- folds == l

# training dataframe
train_df <- data.frame(
  C_DR = C_DR[train, l],
  inc = X[train, "inc"]
)

# omit columns with NA values before regression
# train_df <- na.omit(train_df)

# Plot
ggplot(train_df, aes(x = inc, y = C_DR)) +
geom_point(alpha = 0.2) +
geom_smooth(method = "lm") +
coord_cartesian(ylim = c(0, 60000)) +
labs(title = "Treatment effect vs income")

```


```{r, echo = FALSE}
set.seed(123456789)
# ATE calculations
# ATE
C_DR_av <- rowSums(C_DR, na.rm = TRUE)/(K-1)
ate_hat <- mean(C_DR_av)

# The ATE should be equal to the mean of the CATE
mean(CATE)
se_hat  <- sd(C_DR_av) / sqrt(n)
```
DR-learner
1e run
se_hat
[1] 1123.617
mean(pseudooutcomes)
[1] 7897.361
mean(CATE)
[1] 7888.084

Boven- en ondergrens ATE
[1] 10099.65
[1] 5695.072

2e run
se_hat
[1] 1167.328
mean(pseudooutcomes)
[1] 7878.258
mean(CATE)
[1] 7885.001

Boven- en ondergrens ATE
[1] 10166.22
[1] 5590.296

Stratified run
se_hat
[1] 1142.036
mean(pseudooutcomes)
[1] 7888.663
mean(CATE)
[1] 7868.564

Boven- en ondergrens ATE
[1] 10127.05
[1] 5650.273

```{r, echo = FALSE}
set.seed(123456789)
# 95% confidence intervals
ate_hat + 1.96 * se_hat
ate_hat - 1.96 * se_hat
```

```{r, echo = FALSE}
set.seed(123456789)
hist(CATE, main = "CATE")
```
```{r, echo = FALSE}
set.seed(123456789)
# plot 95% CI's
pension$CATE <- CATE
ggplot(pension, aes(x = inc, y = CATE))  +
  geom_smooth() +
  labs(title = "Treatment effect vs income")
```

```{r, echo = FALSE}
set.seed(123456789)
ggplot(pension, aes(x = inc, y = CATE)) +
  geom_point(alpha = 0.2) +
  geom_smooth() +
  labs(title = "Treatment effect vs income")
```

```{r, echo = FALSE}
set.seed(123456789)
ggplot(pension, aes(x = inc, y = CATE)) +
  geom_point(alpha = 0.2) +
  geom_smooth() +
  labs(title = "Treatment effect vs income")
```

```{r, echo = FALSE}
set.seed(123456789)
hist(rowSums(Y1_hat, na.rm = TRUE)/(K-1), main = "Expected responses with treatment", xlab = "E[Y1]")
hist(rowSums(Y0_hat, na.rm = TRUE)/(K-1), main = "Expected responses without treatment", xlab = "E[Y0]")
hist(rowSums(A_hat, na.rm = TRUE)/(K-1), main = "Propensity scores", xlab = "Propensity-score")
```

Here, there is a choice of two orthogonal learners, namely the DR-learner or the R-learner. An advantage of the DR-learner is that it is doubly robust, meaning that if either the outcome or the propensity model is correctly specified, it gives consistent estimates. However, a disadvantage of the DR-learner is that it is unstable in the presence of local overlap failures. This means that for regions of the covariates, the propensity score is close to 0 or 1 (so all individuals are either treated/untreated). Thus, here it is difficult to compare treated and untreated individuals with  the same covariates. This can in turn destablize the DR-learner and lead to an extremely high variance.

To assess whether this forms a problem, we compute a plot with the distributions of the propensity scores seperately for treated/untreated individuals. If these distributions overlap, then treated and untreated individuals are comparable across propensity score values. However, if there are regions where only treated/untreated are observed, then it is preferred to use the R-learner over the DR-learner.


We estimated the conditional average treatment effect (CATE) of eligibility for a 401(k) plan on net financial assets using a orthogonal doubly robust (DR) learner. The nuisance components — the propensity score and the conditional outcome models — were estimated using random forests, using all covariates n order to improve confounding adjustment. In the final-stage the doubly robust pseudo-outcomes were modelled on income only, using a ordinary linear model.

Compared to the earlier causal forest approach, the chosen DR-learner introduces an additional modeling assumption in the final stage. Specifically, by regressing the pseudo-outcomes linearly on income, the method assumes that the treatment effect varies smoothly and approximately linearly with income. In contrast, causal forests estimate treatment heterogeneity fully nonparametrically and therefore allow for more flexible nonlinear patterns and interactions. The advantage of using in the final stage a linear model instead of a Causal Forest is improved interpretability and easier visualization of the treatment effect as a function of income.



Q4 met R-learner
```{r, echo = FALSE}
set.seed(123456789)
# Q4
# R-learner implementation
# Assumptions
# 1) 

# Reference: solutions of lab Causal Prediction
# Other method: https://rdrr.io/github/xnie/rlearner/man/rboost.html

K <- 6 # set to 6 -> 6-fold CF for outer loop and 5-fold DF for inner loop

variables <- c("net_tfa", "e401", "age", "inc", "fsize", "educ", 
          "marr", "twoearn", "db", "pira", "hown")

Y <- pension$net_tfa
A <- pension$e401
X <- pension[, c("age", "inc", "fsize", "educ", 
                                 "marr", "twoearn", "db", "pira", "hown")]
AX <- pension[, c("e401", "age", "inc", "fsize", "educ", 
                               "marr", "twoearn", "db", "pira", "hown")]
n <- nrow(X)
eps <- 0.01 # epsilon for truncation of propensity-scores

Yx_hat <- matrix(NA, nrow = n, ncol = K)
A_hat <- matrix(NA, nrow = n, ncol = K)
C_R <- matrix(NA, nrow = n, ncol = K)
W_R <- matrix(NA, nrow = n, ncol = K)
CATE <- rep(NA, n)
```


Q4 met R-learner
```{r, echo = FALSE}
set.seed(123456789)
# Q4
# Random folds
# folds <- sample(rep(1:K, length.out = n))

# Stratified folds
folds <- rep(NA, n)

# indices per class
idx0 <- which(A == 0)
idx1 <- which(A == 1)

# random sampling within classes
idx0 <- sample(idx0)
idx1 <- sample(idx1)

# folds
folds[idx0] <- rep(1:K, length.out = length(idx0))
folds[idx1] <- rep(1:K, length.out = length(idx1))
```

```{r, echo = FALSE}
set.seed(123456789)
# Q4
# R-learner implementation
# Loop

for (l in 1:K) {
  
  k_range <- 1:K
  k_range <- k_range[k_range != l]
  
  for (k in k_range) {
    train_k <- folds %in% k_range & folds != k
    test_k <- folds == k
    
    # E[Y|X]
    rf_Yx <- regression_forest(X[train_k, ], Y[train_k])
    Yx_hat[test_k, l] <- predict(rf_Yx, X[test_k, ])$predictions
    
    # E[A|X]
    rf_A <- regression_forest(X[train_k, ], A[train_k])
    A_hat_temp <- predict(rf_A, X[test_k, ])$predictions
    A_hat[test_k, l] <- A_hat_temp
  }
  
  train <- folds != l
  test <- folds == l
  
  # Prevent problems with dividing to zero
  denom <- A[train] - A_hat[train, l]
  eps <- 1e-6
  denom[abs(denom) < eps] <- eps

  # pseudooutcomes C
  C_R[train, l] <- (Y[train] - Yx_hat[train, l]) / denom
    
  # weights for regression of the pseudooutcomes in the combined training data set using weighted linear model
  W_R[train, l] <- denom^2
    
  # training dataframe
  train_df <- data.frame(
    C_R = C_R[train, l],
    W_R = W_R[train, l],
    inc = X[train, "inc"]
  )
  
  # omit columns with NA values before regression
  # train_df <- na.omit(train_df)#
  
  # Regress the pseudooutcomes in the combined training data set
  lm_fit <- lm(C_R ~ inc, data = train_df, weights = W_R, na.action = na.fail)

  # test dataframe
  test_df <- data.frame(
    inc = X[test, "inc"]
  )

  # Predict CATE on the test set
  CATE[test] <- predict(lm_fit, newdata = test_df)
}


```

```{r, echo = FALSE}
set.seed(123456789)
# ATE calculations
# The ATE should be equal to the mean of the CATE
mean(CATE)
se_hat  <- sd(CATE) / sqrt(n)
```

R-learner 1ste run
mean(CATE)
[1] 8049.97
sd(CATE) / sqrt(n)
[1] 44.48384
mean(CATE) + 1.96 * sd(CATE) / sqrt(n)
[1] 8137.159
mean(CATE) - 1.96 * sd(CATE) / sqrt(n)
[1] 7962.782

```{r, echo = FALSE}
set.seed(123456789)
hist(CATE, main = "CATE")
```
```{r, echo = FALSE}
set.seed(123456789)
pension$CATE <- CATE
ggplot(pension, aes(x = inc, y = CATE)) +
  geom_point(alpha = 0.2) +
  geom_smooth() +
  labs(title = "Treatment effect vs income")
```
```{r, echo = FALSE}
set.seed(123456789)
hist(rowSums(Yx_hat, na.rm = TRUE)/(K-1), main = "Expected responses", xlab = "E[Y]")
hist(rowSums(A_hat, na.rm = TRUE)/(K-1), main = "Propensity scores", xlab = "Propensity-score")
```

```{r, echo = FALSE}
set.seed(123456789)
# Q4 Testen for extreme Propensity scores
A_hat_vector <- rowSums(A_hat, na.rm = TRUE)/(K-1)
mean(A_hat_vector < 0.01 | A_hat_vector > 0.99)
mean(A_hat_vector < 0.05 | A_hat_vector > 0.95)
```
mean(A_hat_vector < 0.01 | A_hat_vector > 0.99)
[1] 0
mean(A_hat_vector < 0.05 | A_hat_vector > 0.95)
[1] 0.008976299
=======
---
title: "Causal Machine Learning assignment"
author: "Thomas Sertijn, Bart Smets, Ilja Van Bever"
date: "`r Sys.Date()`"
output: 
pdf_document:
  latex_engine: xelatex
header-includes:
  - \usepackage{tikz}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(hdm)
library(grf)
library(dplyr)
library(tidyr)
library(knitr)
#load data
data(pension)
library(ggdag)
library(ggplot2)
```

```{r, echo = FALSE}
# Cross-fitting
set.seed(4321)
# Reference doc:
# https://grf-labs.github.io/grf/reference/causal_forest.html
# https://grf-labs.github.io/grf/reference/regression_forest.html

variables <- c("net_tfa", "e401", "age", "inc", "fsize", "educ", 
          "marr", "twoearn", "db", "pira", "hown")


Y <- pension$net_tfa
A <- pension$e401
X <- pension[, c("age", "inc", "fsize", "educ", 
                                 "marr", "twoearn", "db", "pira", "hown")]
AX <- pension[, c("e401", "age", "inc", "fsize", "educ", 
                                 "marr", "twoearn", "db", "pira", "hown")]

K <- 5 # set to 5!
n <- nrow(X)
folds <- sample(rep(1:K, length.out = n))
# eps <- 0.05 # epsilon for truncation of propensity-scores

Y_X_hat <- rep(NA, n)
Y_AX_hat <- rep(NA, n)
A_hat <- rep(NA, n)

for (k in 1:K) {
  
  train <- which(folds != k)
  test  <- which(folds == k)
  
  # E[Y|X] met regression forests
  rf_Y_X <- regression_forest(X[train, ], Y[train])
  Y_X_hat[test] <- predict(rf_Y_X, X[test, ])$predictions
  
  # E[A|X]
  rf_A <- regression_forest(X[train, ], A[train])
  A_hat[test] <- predict(rf_A, X[test, ])$predictions
  # truncation propensity scores
  #A_hat[test] <- pmin(pmax(A_hat_temp, eps), 1 - eps)
  #A_hat[test] <- A_hat_temp
  
  # E[Y|A, L] (exercise 3)
  rf_Y_AX <- regression_forest(AX[train, ], Y[train])
  Y_AX_hat[test] <- predict(rf_Y_AX, AX[test, ])$predictions
}
```

```{r, echo = FALSE}
set.seed(4321)

# Propensity scores
pension$A_hat <- A_hat
# 1a) Train Causal Forest
causal_forest <- causal_forest(X = X, Y = Y, W = A,
  Y.hat = Y_X_hat,
  W.hat = A_hat,
  num.trees = 2000,
  min.node.size = 5
)
```

```{r, echo = FALSE}
set.seed(4321)
# Extract CATE and variances
tau.hat.oob <- predict(causal_forest, estimate.variance = TRUE)
pension$CATE <- tau.hat.oob$predictions

pension$CATE_se <- sqrt(tau.hat.oob$variance.estimates)

# 95% betrouwbaarheidsintervallen
pension$CATE_lower <- pension$CATE - 1.96 * pension$CATE_se
pension$CATE_upper <- pension$CATE + 1.96 * pension$CATE_se
```

```{r, echo = FALSE}
set.seed(4321)
# Determining the ATE and his confidence interval
ate <- average_treatment_effect(causal_forest,
                               target.sample = "all",
                               method = "AIPW")

# ATE confidence intervals
ci_low  <- ate[1] - 1.96 * ate[2]
ci_high <- ate[1] + 1.96 * ate[2]
```

# Introduction
In this project the effect of eligibility for enrolling in a 401(k) plan on net financial assets is investigated.

# 1a) Effect estimation with Causal Forests

In the first phase Causal Forests were used to estimate the effect of eligibility for enrolling in a 401(k) plan on net financial assets, in function of age, income, family size, years of education, a married indicator, a two-earner status indicator, a defined benefit pension status indicator, an IRA (individual retirement account) participation indicator, and a home ownership indicator.

The causal forest analysis reveals a causal effect of eligibility for enrollment in a 401(k) plan on accumulated net financial assets. The estimated Average Treatment Effect (ATE) equals `r round(ate[1], 3)`, indicating that individuals who are eligible for a 401(k) plan accumulate, on average, approximately \$`r round(ate[1], 3)` more in net financial assets than non-eligible individuals. The estimated standard error is \$`r round(ate[1], 3)`, resulting in a 95% confidence interval of [\$`r round(ate[1] - 1.96 * ate[2], 3)`, \$`r round(ate[1] + 1.96 * ate[2], 3)`]. Because the confidence interval is positive and larger than 0, we conclude that being eligible for the 401(k) has a significant and positive effect on the net financial assets of a person.

In addition to the average treatment effect, the Conditional Average Treatment Effect (CATE) was examined to investigate heterogeneity in treatment effects across individuals. Figure \@ref(fig:fig-cate-hist) presents a histogram of the predicted CATE values. Most individuals have positive CATE values, although negative values are also observed. Furthermore, the estimated CATE was plotted against income (Figure \@ref(fig:fig-cate-unc)), revealing a clear upward trend: the effect of 401(k) eligibility appears to increase with income. Thus, becoming eligible does have a positive impact on financial assets, but mainly so for richer individuals.

To better characterize this relationship, both a linear regression model and a smoothing spline regression (with $\lambda$ = 1) were fitted. The corresponding trends are displayed in Figures \@ref(fig:fig-cate-inc-smooth) and \@ref(fig:fig-cate-inc-lm) While the estimated curves differ somewhat at higher income levels, this part of the income distribution contains relatively few observations, implying considerably greater uncertainty in this region.

To assess the stability of the inverse probability weights, the distribution of the estimated propensity scores was also investigated using a histogram (Figure \@ref(fig:fig-A-hist)). `r round(mean(A_hat < 0.01 | A_hat > 0.99), 2)`% of the scores were smaller than 0.01 or larger than 0.99, and only `r round(mean(A_hat < 0.05 | A_hat > 0.95), 2)`% of the scores were smaller than 0.05 or larger than 0.95. Overall, the propensity scores do not exhibit problematic extremes. In addition, a clear relationship between income and the propensity scores was observed (Figure \@ref(fig:fig-A-inc)). This finding is inconsistent with the assumption of exchangeability conditional on income alone. Together with the limited number of extreme propensity scores, this provides an additional reason not to apply propensity score truncation. Removing observations with extreme propensity scores would likely shift the estimated ATE toward the treatment effect for higher-income individuals, thereby reducing the representativeness of the overall estimate. This can be viewed as an example of the bias--variance trade-off. Although trimming extreme propensity scores can reduce the variance of the estimator by limiting the influence of highly weighted observations, it may simultaneously introduce additional bias by systematically excluding specific subpopulations from the analysis.

To estimate the propensity scores and outcomes, 5-fold cross-fitting was applied. The modeling was performed using Regression Forests, as described in the documentation of the GRF package (https://grf-labs.github.io/grf/reference/causal_forest.html).

```{r fig-cate-inc-smooth, echo=FALSE, fig.cap="Treatment effect vs income (smoothing spline)"}
set.seed(4321)

# CATE modelled using smoothing spline (lambda=1)
# spline fit
fit <- smooth.spline(x = pension$inc,
                     y = pension$CATE,
                     lambda = 1)

# voorspellingen op een rooster van x-waarden
x_grid <- seq(min(pension$inc),
              max(pension$inc),
              length.out = 500)

pred <- predict(fit, x = x_grid)

# plot
# dataframe maken voor ggplot
pred_df <- data.frame(
  inc = pred$x,
  CATE = pred$y
)

# plot
ggplot(pension, aes(x = inc, y = CATE)) +
  geom_point(alpha = 0.2) +
  geom_line(data = pred_df,
            aes(x = inc, y = CATE),
            color = "blue",
            linewidth = 1.2) +
  labs(title = "Treatment effect vs income")
```

```{r fig-cate-inc-lm, echo=FALSE, fig.cap="Treatment effect vs income (linear model)"}
set.seed(4321)
# CATE modelled using linear model
ggplot(pension, aes(x = inc, y = CATE)) +
  geom_point(alpha = 0.2) +
  geom_smooth(method = "lm",
              se = FALSE,
              color = "blue",
              linewidth = 1.2) +
  labs(title = "Treatment effect vs income")
```

```{r fig-cate-inc-unc, echo = FALSE}
set.seed(4321)
# plot 95% CI's
ggplot(pension, aes(x = inc, y = CATE)) +
  geom_point(alpha = 0.2) +
  geom_smooth(aes(ymin = CATE_lower, ymax = CATE_upper), stat = "smooth") +
  labs(title = "Treatment effect vs income (with uncertainty)")
```

```{r fig-cate-hist, echo=FALSE, fig.cap="Histogram of the predicted CATE values"}
set.seed(4321)
hist(pension$CATE, main = "CATE", xlab = "CATE")
```

```{r fig-A-hist, echo = FALSE}
set.seed(4321)
hist(A_hat, main = "Propensity scores", xlab = "Propensity scores")
```

```{r fig-A-inc, echo = FALSE}
set.seed(4321)

ggplot(pension, aes(x = inc, y = A_hat)) +
  geom_point(alpha = 0.2) +
  geom_smooth() +
  labs(title = "Propensity score vs income")
```

# 1b) Possible problems when the analysis additionally conditions on other variables
It was previously stated that eligibility for a 401(k) could be taken as exchangeable conditional on income. However, conditioning on additional variables can have important implications depending on their role in the causal structure.

## Role of Additional Variables in a Causal Diagram

In addition, including variables in an analysis can have different effects depending on their position in the causal diagram.

- *Collider*  
  In the case of a collider *(Eligibility → Collider ← Outcome)*, conditioning on this variable introduces collider bias.  
  Conditioning on a collider opens a backdoor pathway, making the estimated effect of eligibility biased and very difficult to interpret.

- *Mediator*  
  In the case of a mediator *(Eligibility → Mediator → Outcome)*, part of the effect of eligibility is captured through an indirect effect (via the mediator).  
  Including a mediator is therefore not incorrect, but it implies that the estimated effect is no longer the total effect, but only the direct effect.

- *Common Cause*  
  In the case of a common cause *(Eligibility ← Common Cause → Outcome)*, it is appropriate to include this variable in the analysis. If a common cause is omitted, a backdoor pathway remains open, making the estimated effect of eligibility not credible. We note that conditional exchangeability given only income does not hold if there exist common causes of Eligibility and the outcome that are not controlled for.

- *Variable Affecting Only Eligibility*  
  If a variable only affects eligibility, including it in the analysis does not introduce bias.

- *Variable Affecting Only the Outcome*  
  If a variable only affects the outcome and not eligibility, it can be included in the analysis.  
  Including such variables may improve the efficiency of the estimation.

## Examples

- *Age* could potentially be a *common cause*.  
  Older individuals are generally more likely to be eligible for a 401(k) program (stable full-time jobs, longer job tenure), while older people have had more time to accumulate wealth and financial assets over the course of their lives.

- *IRA* could act as a *collider*, since individuals who are ineligible for a 401(k) may substitute toward an IRA as an alternative retirement savings vehicle, while wealthier individuals with higher net financial assets are also more likely to participate in an IRA.

- *Years of Education* could be a *common cause*.  
  Individuals with higher levels of education tend to accumulate greater net financial assets and are also more likely to be eligible for a 401(k) plan.
  
% --- COLLIDER ---
\begin{center}
\textbf{Collider} \\
\vspace{0.5cm}
\begin{tikzpicture}[>=stealth, node distance=2cm, thick]
    \node (A) at (0,2) {Income (A)};
    \node (Y) at (6,2) {Outcome (Y)};
    \node (Z) at (3,0) {Collider (Z)};
    \draw[->] (A) -- (Y);
    \draw[->] (A) -- (Z);
    \draw[->] (Y) -- (Z);
\end{tikzpicture}
\end{center}

% --- MEDIATOR ---
\begin{center}
\textbf{Mediator} \\
\vspace{0.5cm}
\begin{tikzpicture}[>=stealth, node distance=2cm, thick]
    \node (A) at (0,2)  {Income (A)};
    \node (M) at (3,2)  {Mediator (M)};
    \node (Y) at (6,2)  {Outcome (Y)};
    \draw[->] (A) -- (M);
    \draw[->] (M) -- (Y);
    \draw[->, bend right=30] (A) to (Y);  % direct effect
\end{tikzpicture}
\end{center}

% --- COMMON CAUSE ---
\begin{center}
\textbf{Common Cause} \\
\vspace{0.5cm}
\begin{tikzpicture}[>=stealth, node distance=2cm, thick]
    \node (A) at (0,2)  {Income (A)};
    \node (Y) at (6,2)  {Outcome (Y)};
    \node (C) at (3,4)  {Common Cause (C)};
    \draw[->] (A) -- (Y);
    \draw[->] (C) -- (A);
    \draw[->] (C) -- (Y);
\end{tikzpicture}
\end{center}

# 2) Investigating heterogeneity based on subgroup analysis
Based on the predictions obtained from the causal forest, the data were split into four subgroups to investigate treatment effect heterogeneity.

```{r, echo = FALSE}
set.seed(4321)
#Q2
pension$quartiles <- ntile(pension$CATE, 4)

# Just run the function explicitly for each quartile
Q1_ATE <- average_treatment_effect(causal_forest, subset = (pension$quartiles == 1))
Q2_ATE <- average_treatment_effect(causal_forest, subset = (pension$quartiles == 2))
Q3_ATE <- average_treatment_effect(causal_forest, subset = (pension$quartiles == 3))
Q4_ATE <- average_treatment_effect(causal_forest, subset = (pension$quartiles == 4))


coef_1  <- Q1_ATE["estimate"]
sd_1   <- Q1_ATE["std.err"]

Low_CI_1  <- coef_1 - (1.96 * sd_1)
High_CI_1 <- coef_1 + (1.96 * sd_1)

coef_2  <- Q2_ATE["estimate"]
sd_2   <- Q2_ATE["std.err"]

Low_CI_2  <- coef_2 - (1.96 * sd_2)
High_CI_2 <- coef_2 + (1.96 * sd_2)

coef_3  <- Q3_ATE["estimate"]
sd_3   <- Q3_ATE["std.err"]

Low_CI_3  <- coef_3 - (1.96 * sd_3)
High_CI_3 <- coef_3 + (1.96 * sd_3)

coef_4  <- Q4_ATE["estimate"]
sd_4   <- Q4_ATE["std.err"]

Low_CI_4  <- coef_4 - (1.96 * sd_4)
High_CI_4 <- coef_4 + (1.96 * sd_4)

coef_1
Low_CI_1
High_CI_1

coef_2
Low_CI_2
High_CI_2

coef_3
Low_CI_3
High_CI_3

coef_4
Low_CI_4
High_CI_4

coef_1
coef_2
coef_3
coef_4

cat("ATE estimates by predicted CATE quartile:\n\n")

cat("Q1 (lowest predicted CATE): ",
    round(coef_1, 2),
    " [", round(Low_CI_1, 2), ", ", round(High_CI_1, 2), "]\n", sep = "")

cat("Q2: ",
    round(coef_2, 2),
    " [", round(Low_CI_2, 2), ", ", round(High_CI_2, 2), "]\n", sep = "")

cat("Q3: ",
    round(coef_3, 2),
    " [", round(Low_CI_3, 2), ", ", round(High_CI_3, 2), "]\n", sep = "")

cat("Q4 (highest predicted CATE): ",
    round(coef_4, 2),
    " [", round(Low_CI_4, 2), ", ", round(High_CI_4, 2), "]\n", sep = "")

```

First, we can see that the treatment effect in all four quartiles is positive. This suggests that, on average, 401(k) eligibility has a positive impact on net financial assets across all four quartiles. 

In terms of treatment heterogeneity, we find that the threatment effects for the bottom three quartiles are relatively similar, ranging from $5,826 to $6,437. However, for the 25% of individuals predicted to benefit the most, the treatment effect almost doubles. 

Thus, although the confidence intervals still slightly overlap, still a non-linear treatment effect is suggested. The causal forest did thus succesfully find a sub-population (likely those with higher incomes or higher saving preferences), for which the tax-advantaged saving mechanism is effective.

# 3) Investigating heterogeneity using the Efficient Influence Function
The Efficient Influence Function (EIF) was used to construct a debiased estimator for assessing treatment effect heterogeneity.

```{r, echo = FALSE}
set.seed(4321)
#Q3

## CATE predictions (E(Y1-Y0)/L)
CATEhat<-pension$CATE

## prop score (P(A=1/L))
propscore <- A_hat # propensity scores are already truncated in 1a)

# zonder cross-fitting: prop_score<-causal_forest$W.hat

print(causal_forest)

## E(Y/L)
var <- Y_X_hat
# zonder cross-fitting: var<-causal_forest$Y.hat

#Get actual average treatment effect (E(Y1-Y0))
atehat<-average_treatment_effect(causal_forest)["estimate"]
#Get actual average treatment effect (E(Y1-Y0))
atehat <- as.numeric(average_treatment_effect(causal_forest)["estimate"])
## E(Y/A,L)
Y_AX_hat

# Fill in the formula for the EIF

# (Squared difference CATE and ATE) (Term 1)
term1<-(CATEhat-atehat)^2

# Inverse Probability Weights (Term 2)
term2<-(A/propscore)- ((1-A)/(1-propscore))

# Residuals (Term 3)
term3<-(Y-Y_AX_hat)
  
# Combine to get efficient influence function
sigma_sq_hat<-mean(term1+2*(CATEhat-atehat)*term2*term3)
# estimate sigma^2, mean(EIF)=0 per definition
IF_sigma<- term1 - sigma_sq_hat +2 * (CATEhat - atehat) * term2 * term3
#get standard error of sigma_sq_est
# ss = sample size
ss<-length(IF_sigma)
sigma_sq_se<- sd(IF_sigma)/sqrt(ss)
#get CI
high_ci<-sigma_sq_hat+1.96*sigma_sq_se
low_ci<-sigma_sq_hat-1.96*sigma_sq_se


cat("Estimated Variance (Sigma^2):", sigma_sq_hat, "\n")
cat("95% CI: [", low_ci, ",", high_ci, "]\n")
```

sigma_sq_hat
16169595
sigma_sq_se
41439971
sigma_sq_hat+1.96*sigma_sq_se
[1] 97391939
sigma_sq_hat-1.96*sigma_sq_se
[1] -65052749




```{r, echo = FALSE}
set.seed(4321)
#Extreme influence function diagnostics
summary(IF_sigma)

# Quantiles 
quantile(abs(IF_sigma), probs = c(0.50, 0.90, 0.95, 0.975, 0.99, 0.995, 1))

# Histogram of influence-function values
hist(IF_sigma,
     breaks = 50,
     main = "Estimated influence function for sigma^2",
     xlab = "Influence-function value")

# Histogram of absolute influence-function values
hist(abs(IF_sigma),
     breaks = 50,
     main = "Absolute influence-function values",
     xlab = "|Influence-function value|")

# most influential observations
abs_IF <- abs(IF_sigma)

top_ids <- order(abs_IF, decreasing = TRUE)[1:20]

top_influential <- pension[top_ids, c(
  "net_tfa", "e401", "age", "inc", "fsize", "educ",
  "marr", "twoearn", "db", "pira", "hown"
)]

top_influential$IF_sigma <- IF_sigma[top_ids]
top_influential$abs_IF_sigma <- abs_IF[top_ids]
top_influential$CATE <- CATEhat[top_ids]
top_influential$ATE <- atehat
top_influential$CATE_minus_ATE <- CATEhat[top_ids] - atehat
top_influential$propensity <- propscore[top_ids]
top_influential$ipw_contrast <- term2[top_ids]
top_influential$residual <- term3[top_ids]

top_influential <- top_influential[order(-top_influential$abs_IF_sigma), ]

knitr::kable(top_influential, digits = 2)
diagnostic_components <- data.frame(
  net_tfa = Y,
  A = A,
  CATE = CATEhat,
  CATE_minus_ATE = CATEhat - atehat,
  propensity = propscore,
  ipw_contrast = term2,
  residual = term3,
  term1 = term1,
  correction = 2 * (CATEhat - atehat) * term2 * term3,
  IF_sigma = IF_sigma,
  abs_IF_sigma = abs(IF_sigma)
)

top_components <- diagnostic_components %>%
  arrange(desc(abs_IF_sigma)) %>%
  slice(1:20)

knitr::kable(top_components, digits = 2)

summary(pension$net_tfa)

quantile(pension$net_tfa,
         probs = c(0, 0.5, 0.9, 0.95, 0.99, 0.995, 1))

hist(pension$net_tfa,
     breaks = 50,
     main = "Distribution of net financial assets",
     xlab = "net_tfa")

```
```{r, echo = FALSE}
# Propensity against covariates
p_hat <- A_hat
par(mfrow = c(1, 2))
plot(X$inc, p_hat, pch = 16, cex = 0.3, col = rgb(0,0,0,0.3),
     xlab = "Income", ylab = expression(hat(e)(L)),
     main = "Propensity vs income")
abline(h = c(0.05, 0.95), col = "red", lty = 2)

plot(X$age, p_hat, pch = 16, cex = 0.3, col = rgb(0,0,0,0.3),
     xlab = "Age", ylab = expression(hat(e)(L)),
     main = "Propensity vs age")
abline(h = c(0.05, 0.95), col = "red", lty = 2)
par(mfrow = c(1, 1))

p_hat_trunc <- pmax(0.05, pmin(0.95, p_hat))

plot_df <- as.data.frame(X)
plot_df$p_hat_raw   <- p_hat
plot_df$p_hat_trunc <- p_hat_trunc

plot_long <- plot_df %>%
  pivot_longer(cols = -c(p_hat_raw, p_hat_trunc),
               names_to = "covariate", values_to = "value")

ggplot(plot_long, aes(x = value, y = p_hat_raw)) +
  geom_point(alpha = 0.1, size = 0.3) +
  geom_hline(yintercept = c(0.05, 0.95), col = "red", lty = 2) +
  facet_wrap(~ covariate, scales = "free_x") +
  labs(title = "Raw propensity vs all covariates",
       y = expression(hat(e)(L)), x = "Covariate value")


ggplot(plot_long, aes(x = value, y = p_hat_trunc)) +
  geom_point(alpha = 0.1, size = 0.3) +
  geom_hline(yintercept = c(0.05, 0.95), col = "red", lty = 2) +
  facet_wrap(~ covariate, scales = "free_x") +
  labs(title = "Truncated propensity vs all covariates",
       y = expression(hat(e)(L)), x = "Covariate value")


ggplot(plot_df, aes(x = inc, y = p_hat_raw)) +
  geom_point(alpha = 0.1, size = 0.3) +
  geom_hline(yintercept = c(0.05, 0.95), col = "red", lty = 2) +
  labs(title = "Propensity vs income (raw vs truncated)",
       y = expression(hat(e)(L)), x = "Income") +
  geom_point(aes(y = p_hat_trunc), alpha = 0.1, size = 0.3, col = "blue")
```

(The 95% confidence interval still overlaps with 0, so there is no evidence of effect heterogeneity.)

To assess treatment effect heterogeneity, the variance of the CATE was calculated as given in the assignment. This gives us an idea of the extent as to which the impact of 401(k) eligibility on the net amount of financial assets is different across households with different covariate characteristics. If the effect is close to 0, then we would conclude that the treatment effect is homogenous for households characterized by similar covariates.

More formally, we estimated:

\[
\sigma^2 = Var\{E(Y^1 - Y^0 \mid L)\},
\]

where \(Y^1\) and \(Y^0\) give the outcomes under treatment (eligibility) and control (non-eligibility), respectively, and \(L\) gives the covariates:

\[
\tau(L) = E(Y^1 - Y^0 \mid L),
\]

Thus, we can write:

\[
\sigma^2 = Var\{\tau(L)\}.
\]

If

\[
\psi = E\{\tau(L)\}
\]

gives the average treatment effect. Then,

\[
\sigma^2 = E\left[\{\tau(L)-\psi\}^2\right].
\]

Thus, \(\sigma^2\) measures the variance of the CATE around the ATE. 

Normally, we would just compute the variance around the CATE's. However, CATEs are estimated and therefore, this could lead to incorrect inference. Therefore, the efficient influence function for \(\sigma^2\) is used, denoted by:

\[
\begin{aligned}
\phi(O) 
=& \{\tau(L)-\psi\}^2 - \sigma^2 \\
&+ 2\{\tau(L)-\psi\}
\left\{
\frac{A}{P(A=1\mid L)}
-
\frac{1-A}{P(A=0\mid L)}
\right\}
\{Y - E(Y\mid A,L)\}.
\end{aligned}
\]

The efficient influence function has mean zero at the true value of the parameter, meaning that

\[
E\{\phi(O)\}=0.
\]

This property can be used to construct an estimator for \(\sigma^2\). In practice, we replace the unknown nuisance functions by estimates and solve the empirical analogue of this mean-zero condition:

\[
\frac{1}{n}\sum_{i=1}^{n}\hat{\phi}_i = 0.
\]

Starting from the efficient influence function, this means solving

\[
\frac{1}{n}\sum_{i=1}^{n}
\left[
\{\hat{\tau}(L_i)-\hat{\psi}\}^2 - \hat{\sigma}^2
+
2\{\hat{\tau}(L_i)-\hat{\psi}\}
\left\{
\frac{A_i}{\hat{p}(L_i)}
-
\frac{1-A_i}{1-\hat{p}(L_i)}
\right\}
\{Y_i-\hat{m}(A_i,L_i)\}
\right]
=0.
\]



We replace the (unknown) nuisance functions by empirical estimates

\[
\hat{\tau}(L_i)
\]

for the estimated CATE from the causal forest,

\[
\hat{\psi}
\]

for the estimated average treatment effect,

\[
\hat{p}(L_i) = \hat{P}(A_i=1\mid L_i)
\]

for the estimated propensity score,

\[
\hat{m}(A_i,L_i) = \hat{E}(Y_i\mid A_i,L_i)
\]

for the estimated outcome regression.

This gives the following debiased estimator:

\[
\hat{\sigma}^2 =
\frac{1}{n}\sum_{i=1}^{n}
\left[
\{\hat{\tau}(L_i)-\hat{\psi}\}^2
+
2\{\hat{\tau}(L_i)-\hat{\psi}\}
\left\{
\frac{A_i}{\hat{p}(L_i)}
-
\frac{1-A_i}{1-\hat{p}(L_i)}
\right\}
\{Y_i-\hat{m}(A_i,L_i)\}
\right].
\]

The first term,

\[
\{\hat{\tau}(L_i)-\hat{\psi}\}^2,
\]

gives the estimated variation in the CATE around the ATE. The second term is a debiasing correction based on the residuals from the outcome regression and inverse probability weighting. This correction accounts for potential misspecification in the nuisance functions.

Then, we estimated the influence-function values as

\[
\begin{aligned}
\hat{\phi}_i 
=& \{\hat{\tau}(L_i)-\hat{\psi}\}^2 - \hat{\sigma}^2 \\
&+ 2\{\hat{\tau}(L_i)-\hat{\psi}\}
\left\{
\frac{A_i}{\hat{p}(L_i)}
-
\frac{1-A_i}{1-\hat{p}(L_i)}
\right\}
\{Y_i-\hat{m}(A_i,L_i)\}.
\end{aligned}
\]

The standard error is estimated as:

\[
\widehat{SE}(\hat{\sigma}^2)
=
\frac{sd(\hat{\phi}_i)}{\sqrt{n}},
\]

and a 95% confidence interval was constructed as

\[
\hat{\sigma}^2
\pm
1.96 \times \widehat{SE}(\hat{\sigma}^2).
\]

Ultimately, we inspected the estimated influence-function values and the inverse probability weights as diagnostics, to see if our estimates and CI's were driven by those.

# Main conclusion

Our estimate is negative, approximately −6.68×10^6. As σ^2 denotes the variance, this cannot be negative. However, the debiased estimator is not constrained to be non-negative in finite samples, so a negative estimate can happen if the heterogeneity is small relative to the sampling variability. The 95% confidence interval includes 0, and this thus means that we cannot concldue that there is significant heterogeneity in the impact of being eligible for 401(k) schemes for different households.

As our estimates are rather large, we also took a deeper look at the values for the efficient influence functions. This check indicated that certain observations had large absolute influence function values. More specifically, these values are mostly driven by households with a high amount of net financial assets (and thus large residuals). We further investigated whether the propensity scores of these observations were close to 0 or 1, and this was not the case. Thus, these values were caused by the very large residual values caused by the right-skewed distribution, not by the propensity scores.

Overall, as the confidence intervals overlap with 0, we cannot conclude treatment effect heterogeneity. However, as these confidence intervals remain very broad, it is also difficult to conclude that there is no heterogeneity. Therefore, it is difficult to take conclusions, as inference is impacted by specific high-asset households.


# 4) Estimation of the causal effect using a orthogonal R-learner
The causal forests that were used in the previous phases have the drawback that they learn the causal effect conditional on many variables, making it both difficult to learn with good accuracy and difficult to visualize. In this fourth phase, we therefore estimate the average effect of eligibility for enrollment in a 401(k) plan on net financial assets conditional only on income. For this purpose, an orthogonal R-learner is used, controlling for the same set of covariates as before to ensure valid confounding adjustment.

Before starting on this question, which orthogonal learner that would be used had to be decided. While both DR learner and the R-learner both are theoretically sound to use in this case, it was ultimately decided to implement the R-learner. The R-Learner was chosen because it is more robust to the local overlap failures visible in the propensity score plot. Where p̂ is extreme, the DR-Learner amplifies errors through IPW division and requires explicit clipping, introducing bias. The R-Learner instead naturally downweights those observations via W_R = (Aᵢ − Âᵢ)², requiring no clipping at all.

Q4 met R-learner
```{r, echo = FALSE}
set.seed(4321)
# Q4
# R-learner implementation
# Assumptions
# 1) 

# Reference: solutions of lab Causal Prediction
# Other method: https://rdrr.io/github/xnie/rlearner/man/rboost.html

K <- 6 # set to 6 -> 6-fold CF for outer loop and 5-fold DF for inner loop

variables <- c("net_tfa", "e401", "age", "inc", "fsize", "educ", 
          "marr", "twoearn", "db", "pira", "hown")

Y <- pension$net_tfa
A <- pension$e401
X <- pension[, c("age", "inc", "fsize", "educ", 
                                 "marr", "twoearn", "db", "pira", "hown")]
AX <- pension[, c("e401", "age", "inc", "fsize", "educ", 
                               "marr", "twoearn", "db", "pira", "hown")]
n <- nrow(X)
eps <- 0.01 # epsilon for truncation of propensity-scores

Yx_hat <- matrix(NA, nrow = n, ncol = K)
A_hat <- matrix(NA, nrow = n, ncol = K)
C_R <- matrix(NA, nrow = n, ncol = K)
W_R <- matrix(NA, nrow = n, ncol = K)
CATE <- rep(NA, n)
```


Q4 met R-learner
```{r, echo = FALSE}
set.seed(4321)
# Q4
# Random folds
# folds <- sample(rep(1:K, length.out = n))

# Stratified folds
folds <- rep(NA, n)

# indices per class
idx0 <- which(A == 0)
idx1 <- which(A == 1)

# random sampling within classes
idx0 <- sample(idx0)
idx1 <- sample(idx1)

# folds
folds[idx0] <- rep(1:K, length.out = length(idx0))
folds[idx1] <- rep(1:K, length.out = length(idx1))
```

```{r, echo = FALSE}
set.seed(4321)
# Q4
# R-learner implementation
# Loop

for (l in 1:K) {
  
  k_range <- 1:K
  k_range <- k_range[k_range != l]
  
  for (k in k_range) {
    train_k <- folds %in% k_range & folds != k
    test_k <- folds == k
    
    # E[Y|X]
    rf_Yx <- regression_forest(X[train_k, ], Y[train_k])
    Yx_hat[test_k, l] <- predict(rf_Yx, X[test_k, ])$predictions
    
    # E[A|X]
    rf_A <- regression_forest(X[train_k, ], A[train_k])
    A_hat_temp <- predict(rf_A, X[test_k, ])$predictions
    A_hat[test_k, l] <- A_hat_temp
  }
  
  train <- folds != l
  test <- folds == l
  
  # Prevent problems with dividing to zero
  denom <- A[train] - A_hat[train, l]
  eps <- 1e-6
  denom[abs(denom) < eps] <- eps

  # pseudooutcomes C
  C_R[train, l] <- (Y[train] - Yx_hat[train, l]) / denom
    
  # weights for regression of the pseudooutcomes in the combined training data set using weighted linear model
  W_R[train, l] <- denom^2
    
  # training dataframe
  train_df <- data.frame(
    C_R = C_R[train, l],
    W_R = W_R[train, l],
    inc = X[train, "inc"]
  )
  
  # omit columns with NA values before regression
  # train_df <- na.omit(train_df)#
  
  # Regress the pseudooutcomes in the combined training data set
  lm_fit <- lm(C_R ~ inc, data = train_df, weights = W_R, na.action = na.fail)

  # test dataframe
  test_df <- data.frame(
    inc = X[test, "inc"]
  )

  # Predict CATE on the test set
  CATE[test] <- predict(lm_fit, newdata = test_df)
}


```

```{r, echo = FALSE}
set.seed(4321)
# ATE calculations
# The ATE should be equal to the mean of the CATE
mean(CATE)
se_hat  <- sd(CATE) / sqrt(n)
```

```{r, echo = FALSE}
set.seed(4321)
hist(CATE, main = "CATE")
```
```{r, echo = FALSE}
set.seed(4321)
pension$CATE <- CATE
ggplot(pension, aes(x = inc, y = CATE)) +
  geom_point(alpha = 0.2) +
  geom_smooth() +
  labs(title = "Treatment effect vs income")
```
```{r, echo = FALSE}
set.seed(4321)
hist(rowSums(Yx_hat, na.rm = TRUE)/(K-1), main = "Expected responses", xlab = "E[Y]")
hist(rowSums(A_hat, na.rm = TRUE)/(K-1), main = "Propensity scores", xlab = "Propensity-score")
```

```{r, echo = FALSE}
set.seed(4321)
# Q4 Testen for extreme Propensity scores
A_hat_vector <- rowSums(A_hat, na.rm = TRUE)/(K-1)
mean(A_hat_vector < 0.01 | A_hat_vector > 0.99)
mean(A_hat_vector < 0.05 | A_hat_vector > 0.95)
```

We estimated the CATE of eligibility for a 401(k) plan on net financial assets using an R-Learner. The nuisance components — the conditional outcome model and the propensity score — were estimated using random forests with all covariates to improve confounding adjustment. In the final stage, the R-Learner pseudo-outcomes were modelled on income only using an ordinary weighted linear model.

Compared to the approach used in question 3, where the model gives us the existence of the heterogeneity and attempts to quantify this value, the R-Learner tells how this heterogeneity varies with income, giving a more informative and directly answers the research question about the income-dependent treatment effect.

Compared to the earlier causal forest approach, the chosen R-Learner introduces an additional modelling assumption in the final stage. Specifically, by regressing the pseudo-outcomes linearly on income, the method assumes that the treatment effect varies approximately linearly with income. In contrast, causal forests estimate treatment heterogeneity fully nonparametrically, allowing for more flexible nonlinear patterns and interactions. The advantage of using a linear model in the final stage is improved interpretability and easier visualisation of the treatment effect as a function of income. If instead of linear models, other models would have been used in the approach, then the interpretability of the relationship between income and the treatment effect would have been more difficult, and it would have been more difficult to visualise this relationship. However, using a linear model also introduces the risk of model misspecification if the true relationship is nonlinear, which could lead to biased estimates of the treatment effect heterogeneity. 

>>>>>>> 659163311b35ecde6c13a0541945d64aa71c3b39
